A contextual fear conditioning paradigm in head-fixed mice exploring virtual reality

Version 1

Abstract

Abstract

Contextual fear conditioning is a classical laboratory task that tests associative memory formation and recall. Techniques such as multi-photon microscopy and holographic stimulation offer tremendous opportunities to understand the neural underpinnings of these memories. However, these techniques generally require animals to be head-fixed. There are few paradigms that test contextual fear conditioning in head-fixed mice, and none where the behavioral outcome following fear conditioning is freezing, the most common measure of fear in freely moving animals. To address this gap, we developed a contextual fear conditioning paradigm in head-fixed mice using virtual reality (VR) environments. We designed an apparatus to deliver tail shocks (unconditioned stimulus, US) while mice navigated a VR environment (conditioned stimulus, CS). The acquisition of contextual fear was tested when the mice were reintroduced to the shock-paired VR environment the following day. We tested three different variations of this paradigm and, in all of them, observed an increased conditioned fear response characterized by increased freezing behavior. This was especially prominent during the first trial in the shock-paired VR environment, compared to a neutral environment where the mice received no shocks. Our results demonstrate that head-fixed mice can be fear conditioned in VR, discriminate between a feared and neutral VR context, and display freezing as a conditioned response, similar to freely behaving animals. Furthermore, using a two-photon microscope, we imaged from large populations of hippocampal CA1 neurons before, during, and following contextual fear conditioning. Our findings reconfirmed those from the literature on freely moving animals, showing that CA1 place cells undergo remapping and show narrower place fields following fear conditioning. Our approach offers new opportunities to study the neural mechanisms underlying the formation, recall, and extinction of contextual fear memories. As the head-fixed preparation is compatible with multi-photon microscopy and holographic stimulation, it enables long-term tracking and manipulation of cells throughout distinct memory stages and provides subcellular resolution for investigating axonal, dendritic, and synaptic dynamics in real-time.

Introduction

Large-scale multi-photon imaging and holographic stimulation offers the ability to accurately track and record from the same neuronal populations over extended periods at high spatiotemporal resolution, record from axons, dendrites, and dendritic spines, and stimulate neurons with single-cell precision (Bando et al., 2021; Chaigneau et al., 2016; Dombeck et al., 2010; Dong et al., 2021; Emiliani et al., 2015; Fan et al., 2023; Hattori & Komiyama, 2022; Heer & Sheffield, 2024; Hernandez et al., 2016; Krishnan et al., 2022; Russell et al., 2022; Sheffield & Dombeck, 2015). These techniques facilitate the study of neural circuit dynamics, synaptic and plasticity mechanisms and enable the real-time manipulation of neural activity, all of which are crucial for understanding the mechanisms underlying memory formation and recall in the brain. However, head-fixation of animals, which is generally required to utilize these techniques, imposes constraints on the range of testable behaviors compared to freely moving conditions (Guo et al., 2014; Juczewski et al., 2020). This limitation is particularly evident in the Pavlovian contextual fear conditioning (CFC) task, a laboratory method for testing associative learning, fear memory formation, and recall (M. E. Bouton, 1993; Michael S. Fanselow, 1980, 1990; J. J. Kim & Fanselow, 1992; LeDoux, 2000; Maren, 2001; Maren et al., 2013).

In the traditional CFC paradigm for freely moving animals, an animal is placed in an environment (conditioned stimulus, CS) that is paired with an aversive stimulus, such as a mild foot shock (unconditioned stimulus, US). Animals are then removed from the environment. When reintroduced to the environment, animals exhibit a species-specific conditioned response, such as fearful freezing behavior in rodents, if they successfully associate the shock-paired environment with the aversive stimulus. The CFC paradigm is one of the most basic conditioning procedures for freely moving animals— it’s simple and robust (Michael S. Fanselow, 1980, 1990). Despite its longstanding use in studies involving freely behaving animals to explore questions on learning and memory (Anagnostaras et al., 2001; M. E. Bouton & King, 1983; Mark E. Bouton & Bolles, 1980; Debiec et al., 2002; M. S. Fanselow & Bolles, 1979; Fendt & Fanselow, 1999; Gewirtz et al., 2000; Holland & Bouton, 1999; Jeansok J. Kim & Jung, 2006; Maren et al., 2013; Stephen Maren & Holt, 2000; Pezze & Feldon, 2004), it has been noticeably absent in studies involving head-fixed animals, with some exceptions.

Two prior studies (Lovett-Barron et al., 2014; Rajasethupathy et al., 2015) reported a version of CFC in head-fixed mice with aversive air puffs as the US. These studies reported the conditioned response animals displayed as lick suppression, not freezing. Although lick suppression after fear conditioning has also been observed in freely moving animals (Leaf & Leaf, 1966; Leaf & Muller, 1965), it has been shown to result from increased freezing (Mark E. Bouton & Bolles, 1980). Therefore, relying solely on lick suppression in the head-fixed preparation makes it more challenging to compare with the extensive previous work in freely moving animals, where freezing is measured as the predominant conditioned fear response. Furthermore, the aforementioned head-fixed CFC studies also had the limitation of simultaneously having both a water reward and an aversive stimulus within the context, potentially confounding the results and complicating the interpretation. Hence, we aimed to develop a paradigm where freezing is the conditioned response and without the presence of a reward within the conditioned context, making it directly comparable to freely moving animals.

Here, we describe a contextual fear-conditioning paradigm for head-fixed mice that resulted in freezing as the conditioned response. The context or CS was a virtual reality (VR)-based environment, which head-fixed mice explored by running on a spherical treadmill. We previously demonstrated the feasibility of this approach (Ratigan et al., 2023) and now provide a detailed overview and in-depth analysis of the behavior. Our main aim is to show that VR-based CFC can elicit freezing responses similar to those seen in freely moving animals, while also examining how key parameters affect this response. Though we explored some important variables, this paper does not attempt to comprehensively investigate all parameters necessary for successful VR-based CFC. Additionally, we performed two-photon imaging of a large population of hippocampal CA1 cells during fear conditioning and tracked the same cells during memory recall across days. We analyzed the impact of our paradigm on place cells in the hippocampus and compared these findings to results from CFC in freely moving animals.

Results

A paradigm for contextual fear conditioning in head-fixed mice exploring virtual reality environments

To establish the contextual fear conditioning protocol (Figure 1A-C), head-fixed mice were water-restricted and trained to run on a treadmill and navigate VR environments for water rewards, as we have demonstrated previously (Dong et al., 2021; Krishnan et al., 2022; Krishnan & Sheffield, 2023; Ratigan et al., 2023). Forward movement on the treadmill translates to forward movement in VR, allowing mice to navigate the VR environment. The environments used were 2m-long linear tracks rich in visual cues (Figure 1). Reaching the track’s end triggered a water reward and defined a completed trial/lap. Mice were then virtually teleported back to the track’s start for the next lap. Unlike freely moving mice that tend to explore new environments, head-fixed mice placed on a treadmill do not automatically start exploring VR environments. However, they can be trained through reinforcement learning using water restriction and water rewards (Dombeck et al., 2010; Dong et al., 2021; Krishnan et al., 2022; Ratigan et al., 2023; Sheffield & Dombeck, 2015).

A contextual fear conditioning paradigm for head-fixed mice navigating virtual reality environments.

(A) Experimental setup created with BioRender.com. Mice were head-restrained with their feet resting on a spherical treadmill. Five large monitors surrounded the mice that displayed virtual reality (VR) environments. Movement on the treadmill advanced the VR display, allowing for context exploration. (B) Mice were water-restricted and trained to run laps in the VR for water rewards. VR environments were 2m-long linear tracks. Mice were trained to achieve >3 laps per minute, which took ∼10 to 14 days (Stage 1). Once well-trained, a “tail-coat” was added to their tails (Stage 2), followed by the removal of the water reward the next day (Stage 3). (C) Once the water reward was removed, mice underwent the fear conditioning protocol the following day. On the first experiment day (Day 0), mice spent 10 minutes in the training VR (Familiar VR) and then another 10 minutes in a new VR (CFC VR). After the initial exploration, mice received mild electric shocks on the tail (4-12 shocks, 0.5-1.2mA in amplitude, 1s long). The next day (Day 1), mice were tested for memory recall by placing them in the Familiar VR and the CFC VR for 5 minutes each in a counterbalanced manner. (D) Schematic of the tail-coat used for delivering mild electric shocks to the mouse’s tail. (Top) View from the top (Bottom) side view. Dimensions are provided for a typical 12-week-old male mouse weighing ∼30g before water restriction (see Methods for more details).

Training to navigate VR environments on a treadmill (Stage 1), using water rewards, was achieved in approximately 10-14 days, after which well-trained mice could run several laps per minute. We set a criterion of >3 laps per minute (i.e., a lap velocity of 0.1 - 0.2 m/s) for mice to move on to the next stage (Stage 2). Once the mice reached this stage (Stage 2), the following day, we added a coat to their tails (Figure 1B). We developed a lightweight, wearable conductive apparatus for the mouse’s tail to ensure minimal discomfort for the mouse. We named this apparatus the "tail-coat" (Figure 1D). This tail- coat enabled us to administer mild electric shocks to the mouse’s tail (see Methods).

When the tail-coat was added, some mice reduced their running speed or refused to move. Some recovered the next day, but if their running speed did not reach at least three laps per minute (∼20% of mice failed to meet this criterion), they were not advanced to the next stage. Mice that maintained this running speed advanced to Stage 3. Data from mice that advanced with minimal changes to their lap running behavior after adding the tail-coat are shown in Supplementary Figure 1A.

The following day (Stage 3), mice were placed in the same VR environment with the tail- coat, but the water reward was removed (Figure 1B). Mice that continued to run at least three laps per minute with the tail-coat and without water reward advanced to the experiment stage (Stage 4). Around 40% of the mice dropped out at this stage. High levels of lap running during training were crucial to ensure mice were undeterred by both the tail-coat and the reward removal (Krishnan et al., 2022; Krishnan & Sheffield, 2023). Our criteria for consistent running behavior with minimal pauses throughout these three stages of training ensured that any baseline freezing behavior was minimal, similar to freely moving animals, and allowed us to attribute freezing after CFC to fear conditioning. In summary, our criteria for selecting mice for the experimental stage of contextual fear conditioning required them to run consistently (>3 laps per minute) with 1) rewards, then 2) with a tail-coat, and then 3) without rewards plus the tail-coat. This threshold resulted in about 40% of the mice proceeding to the experimental stage. Once mice qualified for the experimental stage, they were no longer excluded from analysis due to their behavior. We tested three variations of the head-fixed CFC paradigm. In Paradigm 1, the neutral environment was the training VR environment, and a new VR environment was paired with shocks. In Paradigms 2 and 3, both the neutral and shock-paired environments were new and distinct from the training environment to better control for environmental familiarity. In Paradigms 1 and 2, the tail-coat was removed on the days following CFC, whereas in Paradigm 3, it remained on the mouse on all post-CFC days. With these paradigms, we varied some parameters that may influence fear memory recall, contextual discrimination, fear extinction, and their associated behaviors in head-fixed VR-based CFC paradigms. We detail these variations in the following sections, starting with Paradigm 1.

When tested for contextual fear memory recall, mice displayed increased freezing behavior in the shocked-paired VR environment

On Day 0 of the experimental stage in Paradigm 1 (CFC day), mice (n = 18) were exposed to the same VR environment we used for training, referred to as the Familiar VR from here onwards (Figure 1C), for 10 minutes. We then switched them to a new VR environment (CFC VR). Following baseline exploration of the CFC VR for 10 minutes, we repeatedly applied mild-electric shocks (4-12 shocks, 0.5-1mA in amplitude, 1s long) to the mouse’s tail. We call the period when shocks were delivered “during CFC”. Shocks were pseudo-randomly administered with respect to the animal’s position in the CFC VR, with an inter-shock interval (ISI) of 1 minute. To ensure that the shocks were associated with the context and not a specific position in the environment, no visual cues were explicitly paired with the shock. We quantified the animal’s response to the shocks and observed an immediate increase in running speed following each shock (Supplementary Figure 1B). This was consistent across all mice and shocks and was a reliable indicator that the apparatus was working correctly. Sixty seconds after the last shock was administered, the VR screens were switched off, the tail-coat was removed, and mice were returned to their home cages. The next day (Fear memory recall, Day 1), we tested for recall of the learned association between the VR environment and the fearful stimulus. Mice were exposed to both VRs (familiar VR and CFC VR) for five minutes each in a counterbalanced manner. Water reward remained absent on recall days; however, in this paradigm, the tail-coat was no longer present.

Before fear conditioning, mice exhibited similar behavior in both the Familiar and CFC VRs. We evaluated the time taken to complete a lap and the freezing behavior (defined as time points when the animal’s instantaneous velocity dropped below 1cm/s). Neither metric revealed a significant difference between the two VRs before shock administration (Figure 2A, Supplementary Figure 1D-E). However, after CFC, when the animals were reintroduced to the VRs on recall Day 1, we observed a significant increase in freezing and time taken to complete a lap in the CFC VR compared to the Familiar VR (Figure 2A-D). As soon as mice transitioned to the CFC VR (either from dark or after the Familiar VR) on the recall day, we observed behaviors such as slowing down, freezing, moving backward, and hesitation to move forward (Supplementary Videos 1-2). This response was markedly different from their behavior in the Familiar VR. This difference in behavior can be observed clearly in most mice on the very first lap (Figure 2A-B). Since we removed the reward, we didn’t find licking behavior to be indicative of recall behavior (Supplementary Figure 1C). However, some mice displayed licking behavior in the Familiar VR on Day 0 before fear conditioning, which was absent on Day 1. This is likely because mice had previously received a water reward in the Familiar VR. We will address this caveat in Paradigms 2 and 3.

Head-fixed mice show increased freezing behavior following contextual fear conditioning in the CFC VR.

(A) A single example mouse’s lap running behavior on experiment days, Day 0 and Day 1. Behavior is shown for approximately 3 minutes in all sessions except during CFC, which is shown for the 6 minutes that the session lasted. Frames where freezing was detected (instantaneous velocity <1cm/s) are marked with black dots. The traces on the right show that this mouse increased freezing, decreased velocity, and moved backward in the CFC VR (red traces) but not in the Familiar VR (blue traces) on Recall Day. This mouse received six shocks at 1mA intensity at 60 s inter-stimulus interval (ISI). (B) First two minutes of recall behavior in more mice (n = 5) in Familiar VR versus CFC VR. (C-D) (Left Average freezing percentage on recall day in the very first lap (C) and all laps (D) during the 5 minutes that mice explored the Familiar (blue) and CFC (red) VR. Freezing (%) was calculated as the number of frames where freezing was detected in a lap divided by the total number of frames in each lap. (Right) Delta calculated as the difference in the amount of freezing in the CFC VR compared to the Familiar VR before CFC (Day 0) and after CFC (Day 1). The dashed line represents 0. (E-F) Same as C-D but for the time taken to complete the first lap (E) and all laps (F). Mice displayed an increase in freezing and in time taken to complete a lap in the first lap and, on average, in the CFC VR. In C-F, circles and pluses represent individual mice (n = 18, 16 male and two female mice). In C-F, data was pooled from mice receiving different numbers of shocks (4, 6, 12) at varying intensities (0.5 mA, 0.8 mA, 1 mA, and 1.2 mA), which is separately displayed in Supplementary Figure 2. Lines join data from the same mouse. P-values were calculated using a paired t-test.

When we quantified the freezing behavior in mice, we found it to be, on average, high throughout the five minutes that mice spent in the CFC VR on recall Day 1 (Figure 2C-D, average freezing (%), mean ± 95% Confidence Intervals (CI), Familiar VR: 28.95 ± 5.47, CFC VR: 44.13 ± 7.75, p=0.001, Paired t-test). We also observed a significant increase in freezing during the first lap in the CFC VR compared to the familiar VR (Figure 2B-C, average first lap freezing (%), Familiar VR: 24.13 ± 5.59, CFC VR: 43.71 ± 6.49, p<0.001, Paired t-test). This freezing behavior also translated to mice taking longer to complete a lap in the CFC VR compared to the Familiar VR, both in the first lap (average time taken to complete first lap (s), Familiar VR: 14.90 ± 5.27, CFC VR: 42.78 ± 12.17, p<0.001, Paired t-test) and on average (Figure 2E-F, the average time taken to complete all laps (s), Familiar VR: 17.58 ± 4.18, CFC VR: 42.92 ± 13.96, p=0.001, Paired t-test). The increase in freezing and time taken to complete a lap in the CFC VR compared to the Familiar VR was significantly higher on Day 1 than Day 0, indicating that it results from fear conditioning (Figure 2C-F, p<0.05, Paired t-test).

Based on studies in freely moving animals (Michael S. Fanselow, 1980, 1990; Ramanathan et al., 2018; Russo & Parsons, 2021; Trott et al., 2022), we settled on an intensity of 1 mA and six shocks as a reasonable protocol, which we tested on a sizeable cohort of mice (n = 12). We found that freezing at this intensity and shock number was higher on average in the CFC VR; however, mice showed variability in their responses (Supplementary Figure 2A-B). The majority of mice (7/12) displayed appropriate memory recall (Supplementary Figure 2A-B), indicated by more freezing in the CFC VR compared to the Familiar VR (delta > 0), while 3/12 mice froze more in the Familiar VR (delta < 0), and 2/12 showed equal freezing levels in both VRs (delta ≈ 0). Such variability has been regularly documented in freely moving animals and may not be unique to the VR setup used here (Chu et al., 2024; Dos Santos Corrêa et al., 2019; Navarro-Sánchez et al., 2024; Poulos et al., 2016; Russo & Parsons, 2021; Totty et al., 2021). Next, in a small group of mice (n = 6), we investigated whether changing the number of shocks or the shock amplitude would affect the conditioned fear response (Supplementary Figure 2A-B). We found that most shock amplitudes and numbers of shocks increased freezing behavior on average in the CFC VR on recall day (Supplementary Figure 2A-B). The observed variability was within the same range as that of the larger cohort of mice we tested using the 1 mA, 6-shock protocol (Supplementary Figure 2A-B). A more detailed analysis with a larger sample size may be necessary to understand further how variations in shock parameters affect the conditioned response, as has been done in studies with freely moving animals (Baldi et al., 2004; Cordero et al., 1998; Daviu et al., 2024; Poulos et al., 2016; Quinn et al., 2008).

Finally, we examined the extinction of this conditioned response (Supplementary Figure 2C-D). We placed the animals in both the Familiar and the CFC VR for five minutes over four days (Recall Days 1-4) in a counterbalanced manner and tested their recall behavior. Mice displayed the most substantial increase in freezing in the CFC VR on the first day of recall (Supplementary Figure 2C-D, Average freezing (%): Recall Day 1: Familiar VR: 28.95 ± 5.47, CFC VR: 44.13 ± 7.75). Freezing on the first lap in the CFC VR was also most pronounced on Recall Day 1 (First lap freezing (%): Recall Day 1: Familiar VR: 24.13 ± 5.59, CFC VR: 43.71 ± 6.49). Both average (Familiar VR: 27.78 ± 7.86, CFC VR: 33.19 ± 10.22) and first lap freezing (Familiar VR: 20.30 ± 6.24, CFC VR: 25.56 ± 8.03) in the CFC VR decreased on the second day of recall compared to the first recall day, becoming similar to the Familiar VR. While first lap freezing remained similar between the Familiar and the CFC VR on subsequent Recall Days 3 and 4, we observed variability in the average freezing behavior. Average freezing increased significantly again on Recall Day 3 in the CFC VR (Average freezing (%): Recall Day 3: Familiar VR: 29.05 ± 13.32, CFC VR: 43.89 ± 13.10) and was high in both VRs on Recall Day 4 (Recall Day 4: Familiar VR: 37.66 ± 15.77, CFC VR: 38.04 ± 13.42).

To quantify freezing levels more precisely, baseline freezing on Day 0 was compared to recall-day freezing using a Linear Mixed Effects (LME) Model, with recall days as fixed effects and mouse as a random effect (Tables 1-2). In the CFC VR, mice froze significantly more on Recall Days 1, 3, and 4 compared to Day 0. By Recall Day 4, freezing levels were elevated in both the Familiar and CFC VRs relative to baseline (Table 2). On the other hand, first-lap freezing showed a significant increase only on Recall Day 1 in the CFC VR and remained similar to Day 0 on Recall Days 2-4 in both VRs (Table 1). This pattern suggests that freezing in the CFC VR peaks on the first recall day, with a strong conditioned fear response to the shock-paired VR observed on Recall Day 1. The marked difference in freezing behavior between the CFC VR and Familiar VR during the first lap diminishes after the first recall day. On the other hand, average freezing behavior becomes more variable in subsequent recall days, suggesting that complete fear extinction in Paradigm 1 may take longer than four days. The increase in freezing on Recall Day 4 in the Familiar VR may reflect either fear generalization or a response to the extended lack of reward in a previously rewarded environment. Further research is needed to determine the exact conditions required for complete fear extinction in this paradigm.

In summary, these results demonstrate that head-fixed mice can acquire fear associations in VR environments. Most mice displayed freezing as a conditioned response specifically to the shock-paired VR—behavior that mirrors that of freely moving animals after fear conditioning. However, complete fear extinction in this paradigm may require more than four days of testing.

A second paradigm where head-fixed mice discriminate between two novel VRs

As previously mentioned, the incomplete extinction of licking behavior in the familiar VR before fear conditioning or the gradual increases in freezing behavior in the familiar VR during recall tests may indicate that mice still associate the VR with water rewards, which could be a caveat. Furthermore, in traditional CFC paradigms with freely moving mice, fear behavior has been assessed mainly by comparing two novel contexts, one associated with shocks and the other acting as a control. To address this, we tested another paradigm (Paradigm 2), using a novel VR instead of the familiar VR to act as the control (Figure 3A). One novel VR was assigned as the Control VR, and a second novel VR was assigned as the CFC VR, which would later become paired with the shocks. In this paradigm, we also added an extra habituation day to increase the animal’s pre- exposure time to the novel VRs prior to conditioning (Day -1). Pre-exposure to the context before fear conditioning can enhance contextual fear, as seen in freely moving animals (Biedenkapp & Rudy, 2007; Michael S. Fanselow, 1980, 1990; Maren et al., 2013; Rudy et al., 2002; Wiltgen & Silva, 2007).

Using a novel VR as the neutral environment instead of a familiar VR in a second paradigm results in modest increases in freezing.

(A) The training paradigm is similar to paradigm 1. On experiment day, mice were introduced to two novel VRs, one of which would be associated with the shock (CFC VR) and the other wouldn’t (Control VR). In this paradigm, there was an added habituation day, where mice were exposed to the two VRs for ten minutes. The next day, mice ran in the two VRs again before receiving mild electric shocks in the CFC VR. The recall test occurred the next day. In this cohort, all mice received six shocks at 1 mA intensity. (B-D) These panels show the lap running behavior of a single mouse on experiment days 1, 2, and 3. Behavior is displayed for about 3 minutes in all sessions except during CFC. Frames with freezing detected by a threshold (instantaneous velocity <1cm/s) are marked by black dots. In (D), this mouse shows an increase in freezing, a decrease in velocity, and some backward movement when in the CFC VR (red traces). (E) Comparison between freezing in the first lap (left) and all laps (right) in paradigm 1 (blue and red) versus paradigm 2 (green and red). In paradigm 1, only animals that received six shocks at 1 mA intensity are included. n = 12 mice were used in both paradigms (10 male and two female mice). The scale bar for panels B-D is indicated in B. P-values were calculated using a paired t-test.

The training was done similarly to the previous paradigm. After mice showed consistent running behavior in the Familiar VR with the tail-coat and without reward, they were advanced to the experimental stage. On Day -1, mice were exposed to the two novel VRs for ten minutes each. On Day 0, they were re-exposed to the same two VRs for ten minutes before receiving six one-second-long mild electric shocks (at 1 mA and 1 minute ISI) in one of them - the CFC VR. On Day 1, mice were exposed to the two VRs for five minutes to evaluate memory recall (Figure 3A-B).

Compared to Paradigm 1, in the novel VRs, mice showed no differences in licking behavior across all experimental sessions (Supplementary Figure 3A). There was also no significant difference in freezing behavior and time taken to complete a lap between the two VRs on Day -1 and Day 0 before fear conditioning (Supplementary Figure 3B-C).

Head-fixed mice show heightened fear discrimination during the first lap when discriminating between two novel VRs

We compared the freezing behavior between Paradigm 1 and Paradigm 2 on Recall Day 1 to the same intensity and number of shocks (1 mA, 1 s long, 60 s ISI, six shocks, n=12 mice each paradigm). We observed only modest increases in average freezing behavior and time taken to complete a lap in the CFC VR in Paradigm 2, compared to Paradigm 1 (Figure 3C-D, 4A-E). The effect sizes for the difference in freezing behavior between CFC VR and Control VR in Paradigm 2 were smaller than those between Familiar VR and CFC VR in Paradigm 1 (Cohen’s d: First lap freezing: Paradigm 1: 1.377, Paradigm 2: 0.917. Average Freezing: Paradigm 1: 0.893, Paradigm 2: 0.076). When we quantified freezing by lap on Recall Day 1 in Paradigm 2, we found that after the first lap, mice showed similar freezing in both the Control VR and the CFC VR across all subsequent laps (Supplementary Figure 4). In contrast, in Paradigm 1, mice consistently froze more in the CFC VR compared to the Familiar VR throughout Recall Day 1 (Supplementary Figure 4).

Indeed, when we quantified freezing behavior in the Control VR and CFC VR in Paradigm 2, we found that the most significant increase in freezing in the CFC VR was observed in the first lap (freezing in first lap %: Control VR: 16.50 ± 3.15 CFC VR: 23.19 ± 5.75, p = 0.026, time taken to complete first lap (s): Control VR: 14.27 ± 2.85, CFC VR: 25.68 ± 5.55, p = 0.001, Paired t-test, Figure 3, Figure 4A-E). Neither average freezing (Figure 4C, average freezing in all laps %: Control VR: 30.32 ± 6.25, CFC VR: 31.06 ± 6.07, p=0.792, Paired t-test) nor time taken to complete all laps (Figure 4E, the average time taken to complete all laps: Control VR: 21.73 ± 8.11, CFC VR: 24.27 ± 6.36, p=0.558, Paired t-test) showed significant differences between the two VRs.

In Paradigm 2, the most significant freezing and reduced speed occurred in the first lap of the CFC VR compared to the Control VR.

(A) First two minutes of recall behavior in five mice in Control VR (green) versus CFC VR (red). (B-C) The left panels show the amount of freezing on the recall day in (B) the very first lap and (C) all laps during the 5 minutes when the mice explored the Control (green) and CFC (red) VR. Freezing (%) is calculated as the number of frames where freezing was detected in a lap divided by the total number of frames in each lap. On the right, the delta is calculated as the difference in the amount of freezing in the CFC VR compared to the Control VR before (Day 2) and after CFC (Day 3). The dashed line represents 0. (D-E) Same as B-C but for the time taken to complete the first lap (D) and all laps (E). In B-E, circles and pluses represent individual mice (n = 12). Lines join data from the same mouse. P-values were calculated using a paired t-test.

When mice were returned to the two VRs the following day for a second day of recall (Recall Day 2), we found no significant difference in first lap freezing between the Control and CFC VR (Supplementary Figure 5, First lap freezing (%) Recall Day 2: Control VR: 13.96 ± 4.24, CFC VR: 21.34 ± 6.79, p = 0.083, Paired t-test). Average freezing in both CFC VR and Control VR returned to baseline levels on Recall Day 1 and remained so on Recall Day 2 (Supplementary Figure 5, Tables 3-4, Average freezing (%) Recall Day 1: Control VR: 30.32 ± 6.25, CFC VR: 31.06 ± 6.07, p = 0.792, Recall Day 2: Control VR: 33.35 ± 6.11, CFC VR: 35.88 ± 5.89, p = 0.045, Paired t-test). This indicates that fear extinction in Paradigm 2 happens quickly, within the first recall session after the first lap.

A shorter ISI and leaving the tail-coat on during fear memory recall improved fear discrimination when using two novel VRs

Freezing behavior in the CFC VR in Paradigm 2 was weaker than in Paradigm 1, with rapid fear extinction after the first recall lap. In Paradigm 3, two parameters were changed relative to Paradigm 2 to test whether freezing behavior and fear discrimination could be enhanced. First, the interval between consecutive shocks was reduced to see if a shorter ISI could produce a more robust fear response (McNally & Westbrook, 2006; Williams, 1994). Second, the tail-coat was kept on during memory recall, based on evidence that removing contextual cues can reduce conditioned freezing (González et al., 2003). We hypothesized that the tail-coat might act as a contextual cue and that its presence during recall might enhance freezing behavior.

In Paradigm 3, mice followed the same experimental protocol as Paradigm 2, with two novel VRs: one assigned as the Control VR and the other as the CFC VR. A habituation day in the two new VRs (Day -1) was followed by the CFC day (Day 0), where mice received six shocks (0.6 mA, 20-26 s ISI) in the CFC VR. This was followed by recall days (Days 1-3), where mice explored the two VRs presented in a counter-balanced manner. Unlike the other two paradigms, in Paradigm 3, the tail-coat remained on during the recall days (Figure 5A).

A third paradigm that uses a novel VR as the neutral environment but keeps the tail-coat on during memory recall led to increased freezing in the CFC Environment.

(A) The training and experiment paradigm was similar to Paradigm 2 except that the shocks were administered closer together (20-25 s ISI), and the tail-coat was kept on during recall days. (B) First two minutes of recall behavior in Paradigm 3 in five mice in Control VR (green) versus CFC VR (red). (C-D) Comparison between freezing in the first lap (C) and all laps (D) in paradigm 1 (blue and red) versus paradigms 2 and 3 (green and red). n = 12 mice were used in paradigms 1 and 2 (10 male and two female mice), and n = 20 mice in paradigm 3. P-values were calculated using a paired t-test. In Paradigm 3, by keeping the tail-coat on during the recall days, we observed an increase in freezing in the CFC VR in the first lap and across all laps compared to Paradigm 2.

We found that mice (n = 20) showed better fear discrimination between the Control and CFC VRs in Paradigm 3 compared to Paradigm 2 (Figure 5B-D). Mice not only showed more significant freezing in the first lap (freezing in first lap %: Control VR: 17.83 ± 2.95 CFC VR: 25.86 ± 6.51, p = 0.016, Paired t-test) but average freezing across all laps was also higher in the CFC VR compared to the Control VR (Figure 5C-D, Figure 6A-D, average freezing in all laps %: Control VR: 19.86 ± 2.87, CFC VR: 28.19 ± 3.66, p < 0.001, Paired t-test).

Mice froze more and reduced speed in Paradigm 3 across all laps on Recall Day 1; this freezing behavior continued into the second day of recall in Paradigm 3. (A-B)

The right panels show the amount of freezing on the recall day in (A) the very first lap and (B) all laps during the 5 minutes when the mice explored the Control (green) and CFC (red) VR. Freezing (%) is calculated as the number of frames where freezing was detected in a lap divided by the total number of frames in each lap. On the left, the delta is calculated as the difference in the amount of freezing in the CFC VR compared to the Control VR before (Day 2) and after CFC (Day 3). The dashed line represents 0. (C-D) Same as A-B but for time taken to complete the first lap (C) and all laps (D). Circles and pluses represent individual mice (n = 20). Lines join data from the same mouse. P-values were calculated using a paired t-test.

While first-lap freezing showed a weaker effect size in Paradigm 3 compared to the other paradigms, there was a strong effect of average freezing across the session (Cohen’s d: First lap freezing: Paradigm 1: 1.377, Paradigm 2: 0.917, Paradigm 3: 0.744. Average Freezing: Paradigm 1: 0.893, Paradigm 2: 0.076, Paradigm 3: 1.185). On average, we found that Paradigm 1 still produced the highest absolute freezing behavior in the CFC VR (average freezing in all laps %: CFC VR: Paradigm 1: 44.13 ± 7.75, Paradigm 2: 31.06 ± 6.07, Paradigm 3: 28.19 ± 3.66). The difference in freezing behavior between the CFC VR and the neutral VR was greatest in Paradigm 1 (Δ average freezing (CFC VR – Familiar/Control VR): Paradigm 1: 13.14, Paradigm 2: 0.74, Paradigm 3: 8.33). However, unlike Paradigm 2, where the largest differences in freezing between the CFC VR and the Control VR were restricted to the first lap, the shorter ISIs and the addition of the tail- coat during the recall day in Paradigm 3 improved contextual fear discrimination between the two novel VRs. When tested for memory recall in Paradigm 3, mice froze more in the CFC VR than in the Control VR, and this effect persisted beyond the first lap (Figure 5C-D, Supplementary Figure 4).

Additionally, unlike the rapid fear extinction seen in Paradigm 2, mice showed a delay in extinction in Paradigm 3 (Table 5-6, Supplementary Figure 6). Average freezing continued to be higher in the CFC VR than in the Control VR on Recall Day 2 (Average freezing (%) Recall Day 1: Control VR: 19.86 ± 2.87, CFC VR: 28.19 ± 3.66, p < 0.001, Recall Day 2: Control VR: 21.79 ± 3.70, CFC VR: 26.46 ± 3.34, p < 0.001, Paired t-test). By Recall Day 3, there was no significant difference in freezing between the Control VR and the CFC VR (First lap freezing (%): Recall Day 3: Control VR: 13.10 ± 3.37, CFC VR: 17.37 ± 5.15, p = 0.151, Average freezing (%): Recall Day 3: Control VR: 19.43 ± 3.18, CFC VR: 20.73 ± 3.61, p = 0.240, Paired t-test).

However, when comparing average freezing to baseline levels on Day 0 using an LME model, a significant increase in average freezing was observed in both the Control VR and CFC VR groups. This increase persisted across all three recall days (Tables 5-6). Unlike Paradigm 1, in Paradigm 3, freezing levels showed a downward trend over time. However, like Paradigm 1, they had not returned to baseline by the third recall day, suggesting that fear was not entirely extinguished. The presence of the tail-coat during recall days likely contributed to some degree of fear generalization, as mice showed elevated freezing relative to baseline even in the Control VR. However, freezing levels in the Control VR remained much lower than in the CFC VR on all days, indicating that mice continued to discriminate between the VR environments (Figure 6, Supplementary Figure 6, Tables 5-6).

There was no difference in first-lap freezing in the Control VR compared to baseline on any of the recall days. First-lap freezing in the CFC VR was still higher than baseline on Recall Day 2; however, it was back to baseline levels by Recall Day 3 (Table 5-6). As we noticed in Paradigm 1, this indicates that pronounced increases in freezing in the CFC VR in the very first lap are most prominent in the earlier days of recall. On the other hand, the increases in average freezing continue to persist and may take longer to extinguish. Thus, a shorter ISI and the presence of the tail-coat during memory recall may enhance conditioned fear in head-fixed mice when discriminating between two novel VRs. The influence of each parameter on its own will need further investigation.

In summary, our results show that mice exhibited greater average freezing behavior in the CFC VR in Paradigms 1 and 3 compared to Paradigm 2. In Paradigm 2, the most noticeable increase in freezing occurred during the very first lap in the CFC VR compared to the Control VR. We found that fear discrimination could be improved with a shorter ISI when presenting tail shocks and when the tail-coat was present during memory recall, as shown in Paradigm 3. Furthermore, in Paradigms 1 and 3, the fear was not wholly extinguished over the days we tested for recall and may take longer. However, fear extinction was rapid in Paradigm 2 and was back to baseline levels within the first day of recall. Thus, we provide three different variations for performing CFC in head-fixed mice that can be used to understand the neural underpinnings of fear memory.

Place cells in the CA1 subregion of the hippocampus displayed remapping and decreased field widths following contextual fear conditioning

The hippocampus plays a critical role in the encoding, consolidation, and recall of memories and is necessary for the expression of contextual fear memory (Anagnostaras et al., 2001; Ji & Maren, 2007; Liu et al., 2012; Maren et al., 2013; Stephen Maren & Holt, 2000; Phillips & LeDoux, 1992; Ramirez et al., 2013; Wiltgen & Silva, 2007). Context is thought to be represented in the hippocampus by place cells that fire at specific spatial locations (place fields) in an environment (O’Keefe & Dostrovsky, 1971). These place cells have been thought to help discriminate between feared and neutral contexts. Previous studies on freely moving animals have shown that fear conditioning can cause place cells to remap (Blair et al., 2023; E. J. Kim et al., 2015; Kinsky et al., 2023; Mamad et al., 2019; Moita et al., 2004; Ormond et al., 2023; Schuette et al., 2020; Wang et al., 2012, 2015; Wu et al., 2017). Remapping is defined by place cells shifting their firing fields, indicative of place cells adapting to incorporate new information into memory (Colgin et al., 2008; Leutgeb et al., 2005; Muller & Kubie, 1987). Additionally, a recent study (Schuette et al., 2020) that performed calcium imaging of hippocampal cells in freely moving mice found that fear conditioning narrows the width of place fields in the feared environment. This suggests that place cells encode the same environment on a finer scale following fear conditioning, potentially enhancing context discrimination and threat avoidance (Schuette et al., 2020).

We replicated these findings in our head-fixed version of CFC in VR. To investigate the effect of fear conditioning on place cell activity in our task, we expressed a calcium indicator, GCaMP6f (Chen et al., 2013), in hippocampal pyramidal cells and, using two- photon microscopy, imaged from the same hippocampal neurons across days during fear conditioning and fear memory recall in Paradigm 1 (Figure 7A-B). We found that a subset of place cells formed before CFC remained stable, while others remapped or changed their preferred firing locations after CFC (see examples in Figure 7C, 7D-G). This is similar to previously reported observations (Blair et al., 2023; Moita et al., 2004; Wang et al., 2012, 2015).

Fear conditioning results in place cell remapping in both Familiar and CFC VR.

(A) A schematic of the procedure for two-photon imaging of large populations of the same CA1 neurons over multiple days, created with BioRender.com (B) An example field of view across the two experiment days. The same field of view was aligned and imaged across days. (C) Examples of a few place cells in the CFC VR across sessions on Day 0 (before CFC, during CFC) and Day 1 (Memory Recall). White lines separate laps in each session. Some place cells maintain stable fields across days, while others remap. (D) Place fields defined on Day 0 in Familiar VR before CFC are plotted across Day 1 during fear memory recall (n = 4 mice). (E) Place fields defined on Day 0 in CFC VR before CFC are plotted across the fear conditioning session and on Day 1 during fear memory recall. (F) On the left is a scatter plot of the center of mass of place fields defined in Familiar VR before CFC on Day 0 (x-axis) compared to their center of mass on Day 1 during fear memory recall (y-axis). On the right is a boxplot of correlation coefficients between mean place fields defined in Familiar VR before CFC on Day 0 and memory recall on Day 1 (Day 0: Day 1). The within- session correlation coefficients serve as control and were calculated between mean place fields in the first half and second half of the Familiar VR before the CFC session. (G) A scatter plot of the center of mass of place fields defined in CFC VR before CFC on Day 0 (x-axis) compared to their center of mass on (right) Day 0 during CFC and (middle) Day 1 during memory recall. On the left is a distribution of correlation coefficients between mean place fields defined in Familiar VR before CFC on Day 0 and during CFC (Day 0: Day 0) and memory recall on Day 1 (Day 0: Day 1). The within-session correlation was calculated between mean place fields in the first half and second half of the CFC VR before the fear conditioning session. Asterisk (*) indicates significant P values (KS test, P < 0.01). Our findings show that place fields present in the before CFC sessions in both Familiar and CFC VR showed significant remapping following fear conditioning.

We found remapping to occur in both the Familiar and the CFC VR to similar extents, as indicated by the lower correlation across sessions/days compared to the within-session correlation (Figure 7F-G). This contradicts previous findings by Moita et al. (2004), who reported more stability in the control environment compared to the CFC environment after fear conditioning. This remapping in both the Familiar and the CFC VR could be due to differences in the CFC paradigm used in that study versus ours and will need further investigation. It could also result from taking the water reward away from a previously rewarded VR, as is the case in Paradigm 1. We have shown before that this makes the place map more likely to drift across days (Krishnan et al., 2022; Krishnan & Sheffield, 2023). Whether these results hold the same for the Control VR in Paradigms 2 and 3 will need further investigation.

Next, we looked at the place cells in each session. Interestingly, we found an increase in the number of place cells in the CFC VR during memory recall, a finding not previously reported (Figure 8A-B). We also found that place cells had narrower fields in the CFC VR during memory recall (Figure 8C), as previously shown (Schuette et al., 2020). We did not find a significant difference in other place cell parameters we examined, such as lap-by-lap reliability (Figure 8D) and out-of-field firing (Figure 8E). Thus, our observations of remapping and narrower place field widths replicate findings from freely moving animals, validating our paradigm’s use to explore further questions using the head-fixed preparation to study the neural underpinnings of contextual fear memories. We add to this field by showing an increase in the proportion of place cells following CFC, a possible neural substrate of fear memory recall.

The widths of place fields in CA1 narrow during memory recall.

(A) Place fields were defined and sorted by track length in each session, pooled from all mice (n = 4 mice). Each cell’s activity was normalized to its peak, and cells were sorted by their center of mass along the track. (B) The percentage of place cells is calculated as the number of place cells divided by the total number of recorded cells. More place cells were identified by our algorithm in the CFC VR on Day 1. (C-E) Parameters of place fields: (C) Width, (D) Place Field (PF) Reliability, and (E) out/in field firing ratio in both VRs on Day 0 and Day 1. The width of the place fields in the CFC VR significantly decreased on Day 1. No other parameters significantly varied across days. P-values were calculated using a paired t-test.

Discussion

We present three paradigms for performing CFC in head-fixed mice, using a conductive "tail-coat" to deliver aversive shocks as mice navigate VR environments on a treadmill.

Our key findings are: 1) Head-fixed mice exhibited freezing behavior in response to an aversive stimulus in VR, consistent with conditioned responses seen in freely moving animals. Freezing was especially pronounced during the first lap in the fear-conditioned environment compared to the neutral environment across all paradigms tested. 2) With mice on a spherical treadmill, additional conditioned responses such as hesitation and backward movement were detected. 3) The choice of neutral environment influenced fear discrimination: mice froze more in the fear-conditioned VR when the neutral environment was familiar rather than novel. Freezing in the fear-conditioned VR could be enhanced by shorter ISIs and the presence of the tail-coat during recall when the neutral environment was novel. 4) Following fear conditioning, place cells in the hippocampal CA1 region showed remapping, increased numbers, and narrower fields, paralleling findings in freely moving animals and offering new insights.

Here, we experimented with three paradigms for fear conditioning in mice, with the main difference being the neutral environment where the mice didn’t receive any shock. In Paradigm 1, the neutral VR was familiar, where the mice had previously received rewards, and the CFC VR was novel. In Paradigms 2 and 3, both the neutral and CFC VRs were similarly novel. In Paradigm 3, we decreased the ISI and kept the tail-coat on the animal when testing for memory recall. Thus, we have tested three different variations for performing CFC in head-fixed mice. On average, mice showed more freezing behavior in the CFC VR compared to the neutral VR. A common feature across all paradigms was pronounced freezing during the very first lap in the CFC VR. While mice in Paradigm 2 extinguished fear rapidly after the first lap, freezing behavior persisted up to 3-4 recall days in Paradigms 1 and 3, to varying degrees. We also observed variability in conditioned responses across mice. Some displayed an opposite pattern, freezing more in the neutral VR or equally in both VRs. This variability in behavior, similar to that observed in freely moving animals (Chu et al., 2024; Dos Santos Corrêa et al., 2019; Navarro-Sánchez et al., 2024; Poulos et al., 2016; Russo & Parsons, 2021; Totty et al., 2021), can be insightful when combined with investigations of neural activity, helping to understand the neural dynamics that contribute to an animal’s ability to learn the association between a context and fearful stimuli. This can reveal the mechanisms by which such memories may generalize or become maladaptive, as seen in post-traumatic stress disorders.

We have shown here an example of using Paradigm 1 to study place cells in the hippocampus and their role in fear memory formation and recall. We recorded from a large group of the same hippocampal CA1 cells across days and demonstrated that place cells remap and narrow their place field widths with fear conditioning. This occurs in head- fixed mice navigating VR environments, similar to freely moving animals. Narrow place fields may suggest sharper spatial tuning to incorporate a salient emotional memory and may aid in context discrimination (Schuette et al., 2020). However, how place cells narrow their widths in response to fear conditioning remains an open question. One possible way to answer this is by imaging CA1 place cell dendrites to understand how they integrate inputs before and after fear conditioning. We can explore what dendritic firing and plasticity mechanisms lead to narrow place field widths. We can then perturb these firing patterns to attribute causality to the narrowing of place field widths in fear discrimination. Our head-fixed paradigm enables the recording of dendrites and the study of plasticity mechanisms (Sheffield et al., 2017; Sheffield & Dombeck, 2015), thus facilitating these investigations.

We recently used Paradigm 3 to understand the role of the thalamic nucleus reuniens- hippocampal pathway (Ratigan et al., 2023). Using a two-photon microscope, we identified neural activity in this pathway by imaging axons from the nucleus reuniens that project to the CA1 with single-axon resolution. We found this pathway necessary for suppressing fear memory retrieval, context discrimination, and fear extinction. The robust freezing response in Paradigm 3 in the fear-conditioned VR, compared to a neutral novel VR, allowed us to align our findings with the existing literature on the role of the nucleus reuniens-hippocampal circuit in fear memory in freely moving animals (Ramanathan & Maren, 2019; Ramanathan et al., 2018; Totty et al., 2023; Troyner et al., 2018; Vetere et al., 2017; Xu & Südhof, 2013). Furthermore, the subcellular resolution achieved through two-photon calcium imaging using the head-fixed preparation allowed us to longitudinally observe and understand the inputs that axons from the reuniens send to the hippocampus during fear memory formation and retrieval, a level of detail that is difficult to achieve with current techniques in freely moving preparations.

In another unpublished study, we used Paradigm 2 and identified co-active neurons between CA1 and CA3 subregions of the hippocampus during fear memory recall. The difference between the first and subsequent laps has helped us understand the progression of neural activity in the hippocampus as mice go from retrieving the fear memory to extinction within one session. Moreover, the variability in response across mice has allowed us to regress the neural activity to the behavior, identifying neural differences between those mice that accurately recall the fear memory and those that do not. These studies showcase the broad applicability of our different CFC paradigms.

In the laboratory, CFC provokes freezing behavior in freely moving rodents, as demonstrated by several studies that span decades (Michael S. Fanselow, 1980, 1990; Fendt & Fanselow, 1999; Maren, 2001; Maren et al., 2013; Phillips & LeDoux, 1992; Tovote et al., 2015). In our study, we found that a mild electric shock applied to the tail could also trigger freezing behavior as the conditioned response in head-fixed mice, making our paradigm more comparable to the freely moving paradigms than previously attempted methods using air-puff stimulation (Lovett-Barron et al., 2014; Rajasethupathy et al., 2015). There are numerous future opportunities to test how various parameters, such as shock amplitude, number of shocks, duration of fear conditioning, the configuration of the Control versus CFC VR, or adding more sensory cues, may influence the animal’s recall and extinction behavior and how it is encoded in the brain (Michael S. Fanselow, 1990; Huckleberry et al., 2016; Ji & Maren, 2007; Maren & Holmes, 2016; Poulos et al., 2016). Moreover, our technique enables us to conduct a more detailed analysis of the animal’s behavior. For example, besides freezing behavior, we have observed a sprinting behavior or a ’flight response’ in response to each shock delivery and instances of fear-related backward movements during memory recall; defensive behaviors that have also been observed in freely moving animals (Chu et al., 2024; Furuyama et al., 2023; Totty et al., 2021; Trott et al., 2022). As our mice are head-fixed, our setup allows easy measure of pupil diameter, respiration, vocalization, heart rate, and body temperature. These have been shown to change with fear conditioning (Dupin et al., 2020; Godsil et al., 2000; Korn et al., 2017; Shionoya et al., 2013; Stiedl & Spiess, 1997).

There are also opportunities to use machine-learning-empowered algorithms to parse out new behavior markers for contextual fear memory, such as gestures or facial expressions (Datta et al., 2019; Mathis & Mathis, 2020; Syeda et al., 2024; Wiltschko et al., 2020). Thus, we can understand the animal’s behavior in a well-controlled environment, which when combined with neural activity recording, can lead to new insights into the neural circuits mediating different defensive responses and emotional states.

Significant insights into memory mechanisms through the CFC task have also come from recent discoveries of memory engrams in the brain (Josselyn et al., 2015; Josselyn & Tonegawa, 2020; Liu et al., 2012; Tonegawa et al., 2018). Engram cells, identified in various brain regions with the help of genetic tagging and optogenetics, play a crucial role in encoding, consolidating, and retrieving specific memories. Typically, these cells are identified based on the expression of the immediate early gene Fos during fear conditioning. Their artificial activation with optogenetics can trigger memory recall, evidenced by increased freezing behavior. Understanding which neurons form these engram cells, how Fos expression is driven during memory formation, how memory information is stored in these cells, how they evolve with learning, and how their activation results in memory recall is an active area of research (Frankland et al., 2024; Mocle et al., 2024; Monasterio et al., 2024; Pettit et al., 2022; Suthard et al., 2024; Uytiepo et al., 2024). Incorporating our paradigm with simultaneous measurements of cellular and subcellular activity can help answer some of these questions. For instance, a recent study (Monasterio et al., 2024) tagged engram cells during fear conditioning while animals were freely moving and then imaged the spontaneous dynamics of these cells using two- photon imaging before and after fear conditioning. The study revealed that cells forming engrams were inherently more active. A limitation of this study was the separate contexts for imaging versus tagging, which can be addressed with our paradigm. Our CFC paradigm allows for direct neural recording of these cells before, during, and after tagging in the same context to examine how activity during memory formation relates to engram cell emergence, population dynamics, and memory recall. Furthermore, holographic stimulation techniques can reactivate specific cell ensembles of interest to help understand the causal relationships between engram cells and memory. Moreover, the hippocampus contains both place cells and engram cells, both of which appear to be involved in contextual memory. However, their relationship is not yet fully understood. Our CFC paradigm offers an opportunity to study both place cells, as we have shown here, and engram cells simultaneously. This can be achieved by incorporating techniques that allow for in vivo labeling of Fos (Pettit et al., 2022). Thus, our head-fixed version of the CFC paradigm opens new avenues for research into the neural underpinnings of fear learning and memory.

Another advantage of our method is that it allows for studying changes in neuronal dynamics in response to an aversive stimulus, such as a shock. Our fields of view during the shock and subsequent flight-like response periods remained stable, and we could correct image movements during these periods with Suite2p (Pachitariu et al., 2017). This is a significant improvement over one-photon and electrophysiological techniques, where abrupt movements such as those induced by an electric shock can change fields of view and reduce the likelihood of recording from the same groups of cells, thereby hindering longitudinal assessments of neural dynamics. In Paradigm 1, where we imaged hippocampal cells during shock, we did not find cells responding to the shock itself, as others have reported (Barth et al., 2023; Mocle et al., 2024; Suthard et al., 2024). We suspect this could be due to the number or duration of shocks administered and our shocks not being associated with a specific cue or location. In future studies, we plan to optimize these parameters to understand better how neural dynamics change in response to an aversive cue.

Finally, we recognize that our head-fixed CFC paradigm has some limitations. Compared to the freely moving CFC, our setup is less naturalistic. We defined different contexts as two visually distinct VR environments while odor, sound, and tactile cues remained constant. This could have made it harder for mice to differentiate between the two contexts, particularly in Paradigm 2. Unlike freely moving animals in CFC studies, our mice were water-restricted throughout the experimental process. Because the animals need to get used to wearing a tailcoat and running in the VRs without a reward, we tend to ’over-train’ the mice. This criterion excludes mice that are less adept at running, limiting the number that go through the process and requiring larger cohorts of mice. Furthermore, despite our best efforts to include equal numbers of male and female mice, only some female mice reached the experimental stage. We believe this is due to the tail-coat weight and the reduced weight of females after water restriction compared to males (see Methods). Although we observed no behavioral differences between the four tested females and the male mice, we cannot guarantee similar results in other female animals without further testing. Studies indicate that different sexes may display varying conditioned responses, as seen in rats (Gruene et al., 2015; S. Maren et al., 1994). In the future, we aim to address these limitations.

In conclusion, we have developed a head-fixed version of the classic CFC paradigm. Our findings demonstrate that mice exhibit freezing behavior when fear-conditioned in a VR environment. Moreover, we observed that hippocampal place cells remap and narrow their fields in response to fear conditioning, mirroring observations in freely moving animals. This innovative paradigm bridges the behavioral task gap for techniques requiring head-fixation, paving the way for new investigations into neural circuits and deepening our understanding of the neural foundations of emotional memory.

Methods

Subjects

All experimental and surgical procedures adhered to the University of Chicago Animal Care and Use Committee guidelines. This study utilized 10-12 week-old C57BL/6J and Tg(Grik4-cre)G32-4Stl/J mice. Following the start of water restriction, the mice were individually housed in a reverse 12-hour light/dark cycle, and behavioral experiments were conducted during the animals’ dark cycle.

Head-fixation surgery

Mice were anesthetized with 1-2% isoflurane and given an intraperitoneal injection of 0.5 mL of saline and a subcutaneous injection of 0.45 mL of Meloxicam (1-2 mg/kg). A small incision was made on the skull to attach a metal head-plate (9.1 mm x 31.7 mm, Atlas Tool and Die Works) to head-restrain the mice. Some mice also had an imaging window implanted above the hippocampus, as previously described (Dong et al., 2021; Krishnan et al., 2022; Sheffield & Dombeck, 2015). Post-surgery mice were housed individually. After a recovery day, they were water-restricted to 0.8-1.0 ml/day.

Behavior training

Head-restrained mice were trained to run on a spherical treadmill to navigate virtual reality (VR) environments similar to the setups previously described (Dong et al., 2021; Krishnan et al., 2022; Sheffield & Dombeck, 2015). Behavioral training began seven days following the water restriction. Briefly, VR environments were created using the Virtual Reality MATLAB Engine, or VIRMEN (Aronov & Tank, 2014), and projected to five screens covering the mouse’s field of view. Mice were head restrained with their limbs resting on a freely rotating Styrofoam wheel, which acted as a treadmill. VR environments were 2 m linear tracks. Mice received a water reward (4 μl) through a waterspout upon completing each traversal of the linear track, considered a lap. This reward was followed by a 1.5- second pause in the VR to allow the mouse to consume the water. The mice were then virtually teleported to the start of the track to begin a new lap. Training sessions lasted 30 minutes per day. Mice were considered well-trained when they could run at >3 laps per minute. We found that this level of training was necessary to ensure that mice would continue to run even without rewards (Krishnan et al., 2022; Krishnan & Sheffield, 2023). Generally, about 60% of mice met our criteria, which typically took around 10-14 days to achieve. Behavior data was collected using a PicoScope Oscilloscope (PICO4824, Pico Technology, v6.13.2).

Tail-coat

A custom-designed "tail-coat" delivered mild electric shocks to a head-fixed mouse’s tail during virtual reality (VR) navigation. A pictorial representation of the tail-coat can be found in Figure 1D. The tail-coat, handmade from wearable conductive fabrics, featured a lightweight design to ensure the mouse could continue running during baseline VR exploration (Table 7). First, the mouse’s tail was measured while anesthetized for head- plate surgeries. A small denim patch cloth (Michael’s) was cut into a rectangle to fit snugly on the mouse’s tail. Marks were made where snap buttons (Michael’s) would be sewn, ensuring they clipped just above the tail. Given that the mouse’s tail narrows away from the body, the buttons closer to the body were placed slightly further apart than those closer to the tip. Typically, the buttons were set 2 cm apart on the narrow end and 2.5 cm apart on the broader end. Care was taken to adjust the placement of the buttons so they weren’t too tight or loose for the mouse. We found that these measurements were generally similar across sexes and for the age group that we tested. After measurements were taken and the cloth cut, small strips of 8mm wide conductive fabric tape (Adafruit) were attached to either side of the cloth, with some tape hanging off the edge. A metal strip (McMaster Carr) of the same width was affixed to this portion of the tape to provide a secure grip for alligator clips used in fear conditioning. Finally, snap buttons were sewn with conductive thread at the marked spots. Clear nail polish was applied to secure the threads in place. The tail-coat creation process should take less than half an hour. Dimensions of the cloth for a typical male mouse, 12 weeks of age and weighing ∼30 g, are shown in Figure 1D. The denim cloth was 4 cm in width and 3.2 cm in length; the conductive tape was 1.2 cm in width and 5.2 cm in length, with 1.5 cm hanging out at each side.

Contextual fear conditioning protocol in head-fixed mice

Once the mice were well-trained and met our criteria of running >3 laps per minute, they were ready for the next stage. The following day, a tail-coat was attached to the mouse’s tail. Mice that maintained consistent running in the VR, with minimal pauses, while wearing the tail-coat advanced to the next stage. For mice startled by the tail-coat, we typically attempted another day with the tail-coat, rewarding them and observing if their running behavior improved to running at least three laps per minute. If not, these mice were not advanced to the next stage. In mice that do advance, on the subsequent day, the tail-coat was kept on, and the water reward was removed. All else remained the same, including the 1.5 s pause at the end, since we wanted animals to be able to distinguish the start and end of a lap. If the mice continued to run as before, they moved on to the experiment stage. From that day forward, the mice no longer received water reinforcement when running in the VR environments. Typically, about 40% of mice made it to the experiment stage. Tail-coats were only added during VR exploration and removed before returning the animals to their home cage. Once the mice advanced to the experiment stage, they were taken through the entire process and not excluded from analysis, even if they showed signs of inconsistent running behavior, which happened with some mice (for instance, see Supplementary Figure 3B-C).

We implemented three distinct paradigms to assess contextual fear discrimination. In the first paradigm, we conducted experiments over two days (Paradigm 1). On Day 0, mice engaged with the same VR in which they were trained, now referred to as the Familiar VR. After 10 minutes, we transitioned them to a new VR for another 10 minutes; this new VR would later be associated with shocks (CFC VR). During this phase, the mice explored both VRs without any rewards while wearing the tail-coat. After the initial exploration in the CFC VR, they received six mild electric shocks, each lasting 1 second with 1-minute intervals in between. The shocks were administered using a shock generator (Colbourn Precision animal shocker, Harvard Bioscience). A TTL pulse was generated using custom-written codes in MATLAB to turn on the shock generator via an Arduino. The shock times were recorded by the PicoScope Oscilloscope (PICO4824, Pico Technology, v6.13.2). We tested different shock intensities ranging from 0.5 to 1.2 mA. After the last shock, mice were removed from the VR and returned to their home cage. The following day (Day 1), we placed the mice in the Familiar VR or CFC VR for ten minutes to test their memory recall. No rewards were delivered during this phase; however, the tail-coat was not added during memory recall. Some mice underwent additional recall tests over subsequent days to assess the time required for fear memory extinction. As a control, some mice underwent the same paradigm but without any shocks.

In the second paradigm (Paradigm 2), the experiment spanned over three days. On Day - 1, mice explored two novel VRs for ten minutes each, establishing a baseline for context exploration. One of these VRs served as a control, while the other, the CFC VR, was where the mice would later receive electric shocks. On Day 0, the mice again spent ten minutes each in the Control and CFC VRs. After this, they received six mild electric shocks, each lasting one second, with a strength of 1 mA and an inter-trial interval of one minute. The mice wore the tail-coat on both Days -1 and 0. On Day 1, the mice were placed in either the Control VR or CFC VR for ten minutes each, in a counterbalanced manner, to test for memory recall. No rewards were given on all the experimental days; however, the tail-coat was not added to the tails on the recall day (Day 1). The third paradigm (Paradigm 3) was similar to Paradigm 2 and used two novel VRs, except that a tail-coat was also added to the mice’s tails during recall tests.

Additionally, a subset of mice in Paradigms 1 and 2 and all mice in Paradigm 3 went through multiple days of recall tests. The subset of mice was randomly chosen, not based on their recall behavior from the previous day. Mice were presented with either the CFC VR or the Control VR in a counterbalanced manner.

All VRs were randomly assigned as either Control or CFC VR to prevent biasing any inherent preferences in mice. However, the Familiar VR was always kept the same. We also used both sexes in our experiments. However, we found female mice less receptive to the "tail-coat" than their male counterparts. We believe this is due to the low average weight of female mice (∼16 g) following water restriction. Female mice of the same age (10-12 weeks) weighed significantly less than males (20-23 g vs 25-30 g) and lost further weight following water restriction. As a result, only four female mice could proceed through and complete all the behavioral tests. Our ultimate goal is to compare behavior and neural activity across sexes, considering sex as a biological variable. To achieve this, we plan to adjust the age and use female mice of appropriate weight to ensure they can run on the treadmill with the tailcoat in the future.

Behavior measurement

Freezing periods were periods during which the animal’s instantaneous velocity was lower than 1 cm/s. Freezing epochs were defined as continuous, uninterrupted freezing periods longer than 1 s. The time taken to complete a lap was calculated as the total time (in seconds) the animal took to run from 0 to 200 cm.

Surgery and two-photon imaging

Stereotaxic surgeries and cannula window implantation above the CA1 subregion of the hippocampus were performed in the same way as previously reported (Krishnan et al., 2022). Briefly, a genetically-encoded calcium indicator, AAV1-CamKII-GCaMP6f (pENN.AAV.CamKII.GCaMP6f.WPRE.SV40 was a gift from James M. Wilson – Addgene viral prep #100834-AAV1; https://www.addgene.org/100834/; RRID: Addgene_100834) was injected (∼50 nL at a depth of 1.25 mm below the surface of the dura) using a beveled glass micropipette leading to GCaMP6f expression in a large population of CA1 pyramidal cells. Mice underwent water restriction (1.0 ml/day) following viral injection and, seven days later, were implanted with a hippocampal window and head plate (Dombeck et al., 2010). Mice were allowed to rest from the surgeries for 3-4 days, after which behavioral training began. Imaging was done using a laser scanning two-photon microscope (Neurolabware). The setup is the same as the one previously described (Krishnan et al., 2022). Time-series images were collected through Scanbox (v4.1, Neurolabware), and the PicoScope Oscilloscope (PICO4824, Pico Technology, v6.13.2) was used to synchronize frame acquisition timing with behavior. At the end of the imaging session on Day -1 or Day 0, a 1-minute time-series movie was collected at a higher magnification and then averaged to aid as a reference frame in finding the same imaging plane on subsequent days. Across-day images collected were motion-corrected using Suite2p (Pachitariu et al., 2017) and aligned using ImageJ (v1.53, NIH). Region of interests were extracted using Suite2p.

Place cell extraction and place field parameters

Place fields were identified as previously described (Dong et al., 2021; Krishnan et al., 2022). Briefly, the 2 m track was divided into 40 position bins (each 5 cm wide). The running behavior of the animal was filtered to exclude periods where the animal was immobile (speed <1 cm/s). Filtering ensured that place cells were defined only during active exploration and not during freezing bouts. We have previously found that filtering versus not filtering yields similar results (Krishnan et al., 2022). Extracted place fields satisfied the following criteria, which were used for all conditions and all mice: 1. The average ΔF/F was greater than 10% above the baseline. 2. The cell displayed significant calcium transients on >30% of laps in the field. 3. The rising phase of the mean transient was located on the track. 4. Their p-value from bootstrapping was <0.05. Multiple place fields within the same cell were treated independently. Parameters of place fields were calculated as described before (Krishnan et al., 2022); this includes the center of mass (COM), place field reliability, out/in-field firing ratio, and place field widths.

Statistics

For related samples, we performed a paired t-test. Multiple comparisons of related samples were corrected with Bonferroni post-hoc. To quantify freezing across days, we employed a linear mixed-effects model with days of recall as the fixed effect and behavior prior to fear conditioning on Day 0 as baseline. The model also included a random effect for mice to account for the nesting of observations within individual subjects. Separate models were used for each VR. Boxplots are plotted to display the entire distribution of the data. The box in the boxplot ranges from the first quartile (25th percentile) to the third quartile (75th percentile), and the box shows the interquartile range (IQR). The line across the box represents the median (50th percentile). The whiskers extend to 1.5*IQR on either side of the box, and anything above this range is defined as an outlier. A two-tailed Kolmogorov-Smirnov (KS) Test was used to compare distributions. When not using boxplots, mean and confidence intervals are displayed.

eLife Assessment

This useful study presents a virtual reality-based contextual fear conditioning paradigm for head-fixed mice. The approach provides a way to perform multiphoton imaging of neural circuits during behaviors that have traditionally been studied in freely moving animals. However, evidence supporting key claims is currently incomplete, particularly regarding elicitation and detection of freezing behaviors, and the impact of the study would be increased by articulating what this initial exploration of parameters offers over existing approaches.

Reviewer #1 (Public review):

The authors set out to develop a contextual fear learning (CFC) paradigm in head-fixed mice that would produce freezing as the conditioned response. Typically, lick suppression is the conditioned response in such designs, but this (1) introduces a potential confounding influence of reward learning on neural assessments of aversion learning and (2) does not easily allow comparison of head-fixed studies with extensive previous work in freely moving animals, which use freezing as the primary conditioned response.

The first part of this study is a report on the development and outcomes of 3 variations of the CFC paradigm in a virtual reality environment. The fundamental design is strong, with head-fixed mice required to run down a linear virtual track to obtain a water reward. Once trained, the water reward is no longer necessary and mice will navigate virtual reality environments. There are rigorous performance criteria to ensure that mice that make it to the experimental stage show very low levels of inactivity prior to fear conditioning. These criteria do result in only 40% of the mice making it to the experimental stage, but high rates of activity in the VR environment are crucial for detecting learning-related freezing. It is possible that further adjustments to the procedure could improve attrition rates.

Paradigm versions 1 and 2 vary the familiarity of the control context while paradigm versions 2 and 3 vary the inter-shock interval. Paradigm version 1 is the most promising, showing the greatest increase in conditioned freezing (~40%) and good discrimination between contexts (delta ~15-20%). Paradigm version 2 showed no clear evidence of learning - average freezing at recall day 1 was not different than pre-shock freezing. First-lap freezing showed a difference, but this single-lap effect is not useful for many of the neural circuit questions for which this paradigm is meant to facilitate. Also, the claim that mice extinguished first-lap freezing after 1 day is weak. Extinction is determined here by the loss of context discrimination, but this was not strong to begin with. First-lap freezing does not appear to be different between Recall Day 1 and 2, but this analysis was not done. Paradigm version 3 has some promise, but the magnitude of the context discrimination is modest (~10% difference in freezing). Thus, further optimization of the VR CFC will be needed to achieve robust learning and extinction. This could include factors not thoroughly tested in this study, including context pre-exposure timing and duration and shock intensity and frequency.

The second part of the study is a validation of the head-fixed CFC VR protocol through the demonstration that fear conditioning leads to the remapping of dorsal CA1 place fields, similar to that observed in freely moving subjects. The results support this aim and largely replicate previous findings in freely moving subjects. One difference from previous work of note is that VR CFC led to the remapping of the control environment, not just the conditioning context. The authors present several possible explanations for this lack of specificity to the shock context, further underscoring the need for further refinement of the CFC protocol before it can be widely applied. While this experiment examined place cell remapping after fear conditioning, it did not attempt to link neural activity to the learned association or freezing behavior.

In summary, this is an important study that sets the initial parameters and neuronal validation needed to establish a head-fixed CFC paradigm that produces freezing behaviors. In the discussion, the authors note the limitations of this study, suggest the next steps in refinement, and point to several future directions using this protocol to significantly advance our understanding of the neural circuits of threat-related learning and behavior.

Reviewer #2 (Public review):

Summary:

In this manuscript, Krishnan et al devised three paradigms to perform contextual fear conditioning in head-fixed mice. Each of the paradigms relied on head-fixed mice running on a treadmill through virtual reality arenas. The authors tested the validity of three versions of the paradigms by using various parameters. As described below, I think there are several issues with the way the paradigms are designed and how the data are interpreted. Moreover, as Paradigm 3 was published previously in a study by the same group, it is unclear to me what this manuscript offers beyond the validations of parameters used for the previous publication. Below, I list my concerns point-by-point, which I believe need to be addressed to strengthen the manuscript.

Major comments

(1) In the analysis using the LME model (Tables 1 and 2), I am left wondering why the mice had increased freezing across recall days as well as increased generalization (increased freezing to the familiar context, where shock was never delivered). Would the authors expect freezing to decrease across recall days, since repeated exposure to the shock context should drive some extinction? This is complicated by the analysis showing that freeing was increased only on retrieval day 1 when analyzing data from the first lap only. Since reward (e.g., motivation to run) is removed during the conditioning and retrieval tests, I wonder if what the authors are observing is related to decreased motivation to perform the task (mice will just sit, immobile, not necessarily freezing per se). I think that these aspects need to be teased out.

(2) Related to point 1, the authors actually point out that these changes could be due to the loss of the water reward. So, in line 304, is it appropriate to call this freezing? I think it will be very important for the authors to exactly define and delineate what they consider as freezing in this task, versus mice just simply sitting around, immobile, and taking a break from performing the task when they realize there is no reward at the end.

(3) In the second paradigm, mice are exposed to both novel and (at the time before conditioning) neutral environments just before fear conditioning. There is a big chance that the mice are 'linking' the memories (Cai et al 2016) of the two contexts such that there is no difference in freezing in the shock context compared to the neutral context, which is what the authors observe (Lines 333-335). The experiment should be repeated such that exposure to the contexts does not occur on the conditioning day.

(4) On lines 360-361, the authors conclude that extinction happens rapidly, within the first lap of the VR trial. To my understanding, that would mean that extinction would happen within the first 5-10 seconds of the test (according to Figure S1E). That seems far too fast for extinction to occur, as this never occurs in freely behaving mice this quickly.

(5) Throughout the different paradigms, the authors are using different shock intensities. This can lead to differences in fear memory encoding as well as in levels of fear memory generalization. I don't think that comparisons can be made across the different paradigms as too many variables (including shock intensity - 0.5/0.6mA can be very different from 1.0 mA) are different. How can the authors pinpoint which works best? Indeed, they find Paradigm 3 'works' better than Paradigm 2 because mice discriminate better between the neutral and shock contexts. This can definitely be driven by decreased generalization from using a 0.6mA shock in Paradigm 3 compared to 1.0 mA shock in Paradigm 2.

(6) There are some differences in the calcium imaging dataset compared to other studies, and the authors should perform additional testing to determine why. This will be integral to validating their head-fixed paradigm(s) and showing they are useful for modeling circuit dynamics/behaviors observed in freely behaving mice. Moreover, the sample size (number of mice) seems low.

(7) It appears that the authors have already published a paper using Paradigm 3 (Ratigan et al 2023). If they already found a paradigm that is published and works, it is unclear to me what the current manuscript offers beyond that initial manuscript.

(8) As written, the manuscript is really difficult to follow with the averages and standard error reported throughout the text. This reporting in the text occurred heterogeneously throughout the text, as sometimes it was reported and other times it was not. Cleaning this reporting up throughout the paper would greatly improve the flow of the text and qualitative description of the results.

Reviewer #3 (Public review):

Summary:

Krishnan et al. present a novel contextual fear conditioning (CFC) paradigm using a virtual reality (VR) apparatus to evaluate whether conditioned context-induced freezing can be elicited in head-fixed mice. By combining this approach with two-photon imaging, the authors aim to provide high-resolution insights into the neural mechanisms underlying learning, memory, and fear. Their experiments demonstrate that head-fixed mice can discriminate between threat and non-threat contexts, exhibit fear-related behavior in VR, and show context-dependent variability during extinction. Supplemental analyses further explore alternative behaviors and the influence of experimental parameters, while hippocampal neuron remapping is tracked throughout the experiments, showcasing the paradigm's potential for studying memory formation and extinction processes.

Strengths:

Methodological Innovation: The integration of a VR-based CFC paradigm with real-time two-photon imaging offers a powerful, high-resolution tool for investigating the neural circuits underlying fear, learning, and memory.

Versatility and Utility: The paradigm provides a controlled and reproducible environment for studying contextual fear learning, addressing challenges associated with freely moving paradigms.

Potential for Broader Applications: By demonstrating hippocampal neuron remapping during fear learning and extinction, the study highlights the paradigm's utility for exploring memory dynamics, providing a strong foundation for future studies in behavioral neuroscience.

Comprehensive Data Presentation: The inclusion of supplemental figures and behavioral analyses (e.g., licking behaviors and variability in extinction) strengthens the manuscript by addressing additional dimensions of the experimental outcomes.

Weaknesses:

Characterization of Freezing Behavior: The evidence supporting freezing behavior as the primary defensive response in VR is unclear. Supplementary videos suggest the observed behaviors may include avoidance-like actions (e.g., backing away or stopping locomotion) rather than true freezing. Additional physiological measurements, such as EMG or heart rate, are necessary to substantiate the claim that freezing is elicited in the paradigm.

Analysis of Extinction: Extinction dynamics are only analyzed through between-group comparisons within each Recall day, without addressing within-group changes in behavior across days. Statistical comparisons within groups would provide a more robust demonstration of extinction processes.

Low Sample Sizes: Paradigm 1 includes conditions with very low sample sizes (N=1-3), limiting the reliability of statistical comparisons regarding the effects of shock number and intensity. Increasing sample sizes or excluding data from mice that do not match the conditions used in Paradigms 2 and 3 would improve the rigor of the analysis.

Potential Confound of Water Reward: The authors critique the use of reward in conjunction with fear conditioning in prior studies but do not fully address the potential confound introduced by using water reward during the training phase in their own paradigm.

Author response:

The following is the authors’ response to the original reviews

We thank all the reviewers for their time and valuable feedback, which helped us improve our manuscript. Based on the comments, we have made several critical changes to the revised manuscript.

(1) We have changed our threshold for detecting freezing epochs from 1 cm/s to 0 cm/s in this revised manuscript. This change allows us to capture periods when animals are completely still on the treadmill, better matching the "true freezing" behavior seen in freely moving set-ups. We have added a new supplementary video (Supplementary Video 2) that better demonstrates the freezing response we observe. All results and figures in the revised manuscript reflect this updated threshold (Figure 2-6, Supplementary Figures 16, Tables 1-6). Our main findings remain robust, demonstrating that freezing serves as a reliable conditioned response in our paradigms, comparable to freely moving animals. Specifically, freezing behavior increased reliably in the fear-conditioned environment following CFC across all paradigms. We have also added data from a no-shock control group (Supplementary Figure 2) which, when compared to the conditioned group, shows that freezing responses in the conditioned group result from fear conditioning rather than immobility. We do observe other avoidance behaviors unique to our treadmill-based task— such as hesitation, backward movement, and slow crawls. These conditioned behaviors are captured through a separate metric: the time taken to complete a lap.

(2) As suggested by the reviewers, we have separately analyzed fear discrimination and extinction dynamics across recall days (Supplementary Figures 2, 5 and 6, Table 1-6). To assess fear discrimination, we use within-group comparisons to evaluate how well animals differentiate between the two VRs across days. For extinction, we use within-VR comparisons to examine freezing dynamics over time. Freezing across recall days is compared to baseline freezing (pre-conditioning) using a Linear Mixed Effects model (Tables 1-6), with recall days as fixed effects and mouse as a random effect, using baseline freezing as the reference.

(3) We have expanded the behavioral dataset in Paradigm 1 to investigate the effect of shock amplitude on the conditioned fear response (Supplementary Figure 2 C-E). Consistent with findings in freely moving animals, our data show that increasing shock intensity from 0.6 mA to 1.0 mA leads to stronger freezing. For the revised manuscript, we specifically increased the sample size in the 0.6 mA group (n = 8) in Paradigm 1, as this intensity is used in Paradigm 3. These additional data demonstrate that combining a lower shock amplitude with shorter inter-shock intervals and retaining the tail-coat during recall can enhance freezing, suggesting that these parameters help compensate for lower shock intensity.

(4) We have added more sample sizes to the imaging dataset (now n = 8, Figures 7-8).

Finally, we acknowledge that many aspects of this paradigm still require optimization. The headfixed CFC paradigm is in its early stages compared to the decades of research dedicated to understanding fear learning parameters in freely moving CFC paradigms. While there are numerous parameters that could be tested—both those identified through our own discussions and those raised by the reviewers—it is not feasible for a single lab to conduct a full evaluation of all the possible factors that could influence CFC in the head-fixed prep. A key limitation is that our approach requires robust navigation behavior in the VR without rewards, which requires weeks of training per mouse. It also necessitates larger sample sizes at the outset as not all animals will make it through our behavioral criteria required for CFC. Another important consideration is scalability. Unlike freely moving CFC paradigms, which allow parallel testing of many animals with minimal pre-training, the VR-CFC setup requires several weeks of behavior training and involves a more complex integration of hardware and software to accurately track behavior in virtual space. The number of VR rigs that can be operated simultaneously in a single lab is often limited, making high-throughput testing more challenging. These factors mean that the testing of a single parameter in a group of animals requires approximately 3–4 months to complete. Despite these constraints, we are committed to continue refining this paradigm over time. With this manuscript, our main aim was to provide a detailed framework, initial parameters, and evidence for conditioned behavior in the head-fixed preparation. By doing so, we hope to facilitate the adoption of this paradigm by researchers interested in studying the neural correlates of learning and memory using multiphoton imaging and stimulation techniques. This approach enables investigations that are not possible in freely moving animals, while the presence of freezing as a conditioned response allows for direct comparisons to the extensive body of work done in freely moving paradigms. Moving forward, we anticipate that optimizing this paradigm and identifying the key parameters that drive learning will be a collaborative, community-led effort.

Public Reviews:

Reviewer #1 (Public review):

The authors set out to develop a contextual fear learning (CFC) paradigm in head-fixed mice that would produce freezing as the conditioned response. Typically, lick suppression is the conditioned response in such designs, but this (1) introduces a potential confounding influence of reward learning on neural assessments of aversion learning and (2) does not easily allow comparison of head-fixed studies with extensive previous work in freely moving animals, which use freezing as the primary conditioned response.

The first part of this study is a report on the development and outcomes of 3 variations of the CFC paradigm in a virtual reality environment. The fundamental design is strong, with headfixed mice required to run down a linear virtual track to obtain a water reward. Once trained, the water reward is no longer necessary and mice will navigate virtual reality environments. There are rigorous performance criteria to ensure that mice that make it to the experimental stage show very low levels of inactivity prior to fear conditioning. These criteria do result in only 40% of the mice making it to the experimental stage, but high rates of activity in the VR environment are crucial for detecting learning-related freezing. It is possible that further adjustments to the procedure could improve attrition rates.

We acknowledge that further adjustments to the procedure could improve attrition rates, and we will continue to work on improving the paradigm.

Paradigm versions 1 and 2 vary the familiarity of the control context while paradigm versions 2 and 3 vary the inter-shock interval. Paradigm version 1 is the most promising, showing the greatest increase in conditioned freezing (~40%) and good discrimination between contexts (delta ~15-20%). Paradigm version 2 showed no clear evidence of learning - average freezing at recall day 1 was not different than pre-shock freezing. First-lap freezing showed a difference, but this single-lap effect is not useful for many of the neural circuit questions for which this paradigm is meant to facilitate. Also, the claim that mice extinguished first-lap freezing after 1 day is weak. Extinction is determined here by the loss of context discrimination, but this was not strong to begin with. First-lap freezing does not appear to be different between Recall Day 1 and 2, but this analysis was not done.

This is an important point. Following reviewer suggestions, we have replotted our figures for all paradigms to show within-VR freezing (see Supplementary Figures 2, 5 and 6) as the appropriate method for quantifying fear extinction across days. Using an LME model (Tables 16), we quantify freezing during recall days against baseline freezing levels measured before fear conditioning within each VR. In Paradigm 2, while some fear discrimination persists across days, extinction does occur rapidly. After the first lap in the CFC VR, we observed no significant differences in freezing compared to the baseline. These results are shown in the revised Supplementary Figure 5, and the revised text is in lines 393-399.

Paradigm version 3 has some promise, but the magnitude of the context discrimination is modest (~10% difference in freezing). Thus, further optimization of the VR CFC will be needed to achieve robust learning and extinction. This could include factors not thoroughly tested in this study, including context pre-exposure timing and duration and shock intensity and frequency.

We acknowledge that many aspects of this paradigm still need optimization, as virtual reality CFC is in its early stages, and we have not explored all of the parameter space. We describe above the reasoning for this. However, for this revised version of the paper we have added new behavioral data (Supplementary Figure 2 C-E) showing that increasing shock intensities from 0.6 mA to 1 mA enhances freezing, both in the first lap and on average. There are of course many other parameters that are likely important, like the ones pointed out here by the reviewer, but exploring the entire parameter space will take many years and will likely require many labs. The purpose of this paper is to show that VR-CFC fundamentally works and is a starting point from which the field can build on. We have now pointed out in the introduction (lines 54-58) and discussion (lines 730-737, 810-814) that there remains significant scope for improving this paradigm and optimizing parameters in the future.

The second part of the study is a validation of the head-fixed CFC VR protocol through the demonstration that fear conditioning leads to the remapping of dorsal CA1 place fields, similar to that observed in freely moving subjects. The results support this aim and largely replicate previous findings in freely moving subjects. One difference from previous work of note is that VR CFC led to the remapping of the control environment, not just the conditioning context. The authors present several possible explanations for this lack of specificity to the shock context, further underscoring the need for further refinement of the CFC protocol before it can be widely applied. While this experiment examined place cell remapping after fear conditioning, it did not attempt to link neural activity to the learned association or freezing behavior.

This is an interesting observation. We think that the remapping observed in the control context likely occurred due to the absence of reward in a previously rewarded environment. Our prior work has demonstrated that removal of reward causes increased remapping (Krishnan et al., 2022, Krishnan and Sheffield, 2023). In other words, the continued presence of reward within an environment stabilizes CA1 place fields. The Moita et al. (2004) paper, which showed remapping only in the fear conditioned context and not in the control context, provided rats with food pellets throughout the experimental session in both the control and conditioned context— likely to increase exploration necessary for identifying place cells. The presence of reward in the Moita et al experiment could explain the minimal remapping observed in their control context compared to our control context which lacked reward. Another possibility could lie in the differences in the intervals between place cell activity recordings in our study and that of Moita et al. While Moita et al. separated their recordings by just one hour, our recordings were separated by a full day, with a sleep period in between. The absence of sleep and the shorter time interval between conditioning and retrieval sessions in their study could explain the minimal remapping observed by Moita et al. compared to our findings. We have now addressed this discrepancy explicitly in lines 596-606.

Although we agree with the reviewer that it would be informative to perform analysis of how neural activity correlates with freezing responses, we think this warrants its own stand-alone manuscript as the neural dynamics and methods to appropriately analyze them are complicated. We are in the midst of analyzing this data further and will present these findings in a separate publication.

In summary, this is an important study that sets the initial parameters and neuronal validation needed to establish a head-fixed CFC paradigm that produces freezing behaviors. In the discussion, the authors note the limitations of this study, suggest the next steps in refinement, and point to several future directions using this protocol to significantly advance our understanding of the neural circuits of threat-related learning and behavior.

Reviewer #2 (Public review):

Summary:

In this manuscript, Krishnan et al devised three paradigms to perform contextual fear conditioning in head-fixed mice. Each of the paradigms relied on head-fixed mice running on a treadmill through virtual reality arenas. The authors tested the validity of three versions of the paradigms by using various parameters. As described below, I think there are several issues with the way the paradigms are designed and how the data are interpreted. Moreover, as Paradigm 3 was published previously in a study by the same group, it is unclear to me what this manuscript offers beyond the validations of parameters used for the previous publication. Below, I list my concerns point-by-point, which I believe need to be addressed to strengthen the manuscript.

Major comments

(1) In the analysis using the LME model (Tables 1 and 2), I am left wondering why the mice had increased freezing across recall days as well as increased generalization (increased freezing to the familiar context, where shock was never delivered). Would the authors expect freezing to decrease across recall days, since repeated exposure to the shock context should drive some extinction? This is complicated by the analysis showing that freeing was increased only on retrieval day 1 when analyzing data from the first lap only. Since reward (e.g., motivation to run) is removed during the conditioning and retrieval tests, I wonder if what the authors are observing is related to decreased motivation to perform the task (mice will just sit, immobile, not necessarily freezing per se). I think that these aspects need to be teased out.

This is an important point and we agree teasing out a lack of motivation versus fearful freezing would be useful. To address the possibility that reduced motivation to run without reward could contribute to the observed freezing behavior, we have now included a no-shock control group in the revised manuscript (n = 7; Supplementary Figure 2A-B, H–I). These control mice experienced the same protocol, including the wearing of a tail coat, but did not receive any shocks. We observed no increases in freezing across days in these controls, confirming that the increased freezing in the Familiar context of our experimental group stems from fear conditioning rather than the removal of reward from a previously rewarded context. If reduced motivation from reward removal were the primary driver, similar freezing patterns would have emerged in the no-shock controls. We have added lines 248-261 in the revised manuscript, discussing this point, and we thank the reviewer for motivating us to do this experiment and analysis.

That said, the precise mechanisms underlying the fear generalization observed in the nonconditioned context—particularly its emergence during later recall days—remain unclear. Studies in freely moving animals have shown that fear memories initially specific to the conditioned context can become generalized with repeated exposures, which may be occurring here (Biedenkapp & Rudy, 2007; Wiltgen & Silva, 2007). Alternatively, it is possible that the combination of fear conditioning and the removal of expected reward contributes to a delayed generalization effect. This may reflect a limitation of our approach, which relies on reward to motivate initial training. As noted by another reviewer, we have now addressed this potential drawback of reward-based training in the discussion (see lines 809-817). Clearly, unique factors specific to the head-fixed VR paradigm may contribute to this phenomenon. Understanding the mechanisms underlying fear generalization in the head-fixed VR CFC paradigm will be a valuable direction for future research.

(2) Related to point 1, the authors actually point out that these changes could be due to the loss of the water reward. So, in line 304, is it appropriate to call this freezing? I think it will be very important for the authors to exactly define and delineate what they consider as freezing in this task, versus mice just simply sitting around, immobile, and taking a break from performing the task when they realize there is no reward at the end.

As noted in point 1 above, we have added a no-shock control group (n = 7; Supplementary Figure 2A-B, H–I) to determine whether the observed freezing was driven by fear conditioning or by reduced motivation to run in the absence of reward. The absence of increased freezing in these controls supports the interpretation that the behavior in the conditioned group is fearrelated. In future studies, incorporating additional physiological measures—such as heart rate monitoring—could further help distinguish fear-related freezing from other forms of immobility.

(3) In the second paradigm, mice are exposed to both novel and (at the time before conditioning) neutral environments just before fear conditioning. There is a big chance that the mice are 'linking' the memories (Cai et al 2016) of the two contexts such that there is no difference in freezing in the shock context compared to the neutral context, which is what the authors observe (Lines 333-335). The experiment should be repeated such that exposure to the contexts does not occur on the conditioning day.

This is an interesting idea. However, if memory linking were driving the observed freezing patterns, we would expect to see similarly reduced fear discrimination across all three paradigms, as mice experience both contexts sequentially in each case. However, this effect appears to be specific to Paradigm 2, suggesting this may be due to other factors. We agree it would be informative to eliminate pre-conditioning exposure to both environments—to assess whether this improves fear discrimination and helps clarify the potential contribution of memory linking. This is something we plan to do in future studies that are beyond the scope of this initial paper on VR-CFC.

(4) On lines 360-361, the authors conclude that extinction happens rapidly, within the first lap of the VR trial. To my understanding, that would mean that extinction would happen within the first 5-10 seconds of the test (according to Figure S1E). That seems far too fast for extinction to occur, as this never occurs in freely behaving mice this quickly.

We agree with the reviewer that extinction in Paradigm 2 appears to occur relatively rapidly.

However, the average time to complete the first lap in the fear-conditioned context in Paradigm 2 is 25.68 ± 5.55 seconds (as stated in line 384), indicating that extinction occurs within approximately the first 30 seconds of context exposure—not within 5–10 seconds. This is specific to Paradigm 2 and does not happen in either of the other paradigms, as shown in Supplementary Figure 4. For clarification, Figure S1E pertains to baseline running in Paradigm 1 and does not apply to Paradigm 2.

As the reviewer points out, even at 30 seconds, extinction seems to be happening more quickly in Paradigm 2 than seen in freely moving setups. This may be due to a key structural difference in our setup. The VR-CFC task is organized into discrete trials, with mice being teleported back to the start after reaching the end of the virtual track. Completing a full lap without receiving a shock could serve as a clear signal that the threat is no longer present within the environment as the completion of a lap means that the animals have surveyed all locations within the environment. This structure could accelerate extinction compared to freely moving setups, where animals take longer to explore their complete environment due to the lack of discrete trials. Although this is true for all our paradigms, the accelerated extinction seen in paradigm 2 versus 1 and 3 may be driven by other factors. As noted by the reviewers, other task parameters—such as context pre-exposure timing, shock intensity, and conditioning duration— are likely to play a role in shaping extinction dynamics. These factors warrant further investigation, and we plan to explore them in future studies to better understand the conditions influencing extinction in the VR-CFC paradigm.

(5) Throughout the different paradigms, the authors are using different shock intensities. This can lead to differences in fear memory encoding as well as in levels of fear memory generalization. I don't think that comparisons can be made across the different paradigms as too many variables (including shock intensity - 0.5/0.6mA can be very different from 1.0 mA) are different. How can the authors pinpoint which works best? Indeed, they find Paradigm 3 'works' better than Paradigm 2 because mice discriminate better between the neutral and shock contexts. This can definitely be driven by decreased generalization from using a 0.6mA shock in Paradigm 3 compared to 1.0 mA shock in Paradigm 2.

The reviewer brings up important points here. We have now added new data evaluating 0.6 mA shocks in Paradigm 1 (Supplementary Figure 2A–E, n=8). These data show that 1.0 mA shocks produced stronger conditioned responses and greater fear discrimination compared to 0.6 mA. Our goal in Paradigm 3 was to begin with a lower shock intensity and assess whether additional modifications—specifically the shorter ISI and retention of the tail-coat during recall—could enhance fear conditioning. Surprisingly, despite the weaker shock intensity, Paradigm 3 resulted in improved discrimination and freezing behavior relative to Paradigm 2. We have now clarified this point in the manuscript (lines 466-470), and we interpret this outcome as evidence that the shorter ISIs and contextual cue continuity (tail-coat) likely play a more significant role in enhancing learning and recall. However, as noted in the text (lines 511-514), further testing is needed to determine the individual contributions of each parameter to successful VR-CFC. Fully optimizing the parameter settings will take additional time and resources, and we aim to continually refine the parameter space in the future, as has been done over the years for freely moving animals.

(6) There are some differences in the calcium imaging dataset compared to other studies, and the authors should perform additional testing to determine why. This will be integral to validating their head-fixed paradigm(s) and showing they are useful for modeling circuit dynamics/behaviors observed in freely behaving mice. Moreover, the sample size (number of mice) seems low.

The one notable difference between our imaging study and that done in freely moving animals is that we observed remapping of place cells in the control context. In contrast, Moita et al. (2004) reported more stable place fields in the control context. A key distinction is that their study included rewards in the control context, which may have contributed to the spatial stability. We now discuss this difference in the manuscript (lines 599-605).

It should be noted that there are many key distinctions among paradigms that study neural activity during fear conditioning in freely moving animals. These include varying exposure times to environments (1–6 days), the time interval between neural activity recordings, and the use of food rewards during the experiment stages in freely moving animals to encourage exploration for place cell identification. Although freely moving paradigms that investigate fear conditioning and place cells are heterogeneous, we were encouraged by the replication of several key findings. This validates VR-based CFC as a viable tool for neural circuit investigations. While future work will include more thorough analyses, our current findings demonstrate the paradigm's effectiveness for modeling circuit dynamics and behavior. We have now expanded our dataset, which includes four additional mice, further corroborating these original findings.

(7) It appears that the authors have already published a paper using Paradigm 3 (Ratigan et al 2023). If they already found a paradigm that is published and works, it is unclear to me what the current manuscript offers beyond that initial manuscript.

The reviewer is correct that we have published a paper using Paradigm 3. However, this manuscript goes beyond that one and provides a much more comprehensive description and fundamental analysis of the behavior and experimental parameters regarding VR-CFC, allowing the research community to adapt our paradigm reproducibly. While Ratigan et al. (2023) offered only a minimal description of behavior and included just Paradigm 3, we present two additional paradigms along with neuronal validation using hippocampal place cells. We have now explicitly stated this in the introduction (lines 50-55).

(8) As written, the manuscript is really difficult to follow with the averages and standard error reported throughout the text. This reporting in the text occurred heterogeneously throughout the text, as sometimes it was reported and other times it was not. Cleaning this reporting up throughout the paper would greatly improve the flow of the text and qualitative description of the results.

We completely agree with this point and have now cleaned up the text, leaving details only in a few places we felt were important.

Reviewer #3 (Public review):

Summary:

Krishnan et al. present a novel contextual fear conditioning (CFC) paradigm using a virtual reality (VR) apparatus to evaluate whether conditioned context-induced freezing can be elicited in head-fixed mice. By combining this approach with two-photon imaging, the authors aim to provide high-resolution insights into the neural mechanisms underlying learning, memory, and fear. Their experiments demonstrate that head-fixed mice can discriminate between threat and non-threat contexts, exhibit fear-related behavior in VR, and show context-dependent variability during extinction. Supplemental analyses further explore alternative behaviors and the influence of experimental parameters, while hippocampal neuron remapping is tracked throughout the experiments, showcasing the paradigm's potential for studying memory formation and extinction processes.

Strengths:

Methodological Innovation: The integration of a VR-based CFC paradigm with real-time twophoton imaging offers a powerful, high-resolution tool for investigating the neural circuits underlying fear, learning, and memory.

Versatility and Utility: The paradigm provides a controlled and reproducible environment for studying contextual fear learning, addressing challenges associated with freely moving paradigms.

Potential for Broader Applications: By demonstrating hippocampal neuron remapping during fear learning and extinction, the study highlights the paradigm's utility for exploring memory dynamics, providing a strong foundation for future studies in behavioral neuroscience.

Comprehensive Data Presentation: The inclusion of supplemental figures and behavioral analyses (e.g., licking behaviors and variability in extinction) strengthens the manuscript by addressing additional dimensions of the experimental outcomes.

Weaknesses:

Characterization of Freezing Behavior: The evidence supporting freezing behavior as the primary defensive response in VR is unclear. Supplementary videos suggest the observed behaviors may include avoidance-like actions (e.g., backing away or stopping locomotion) rather than true freezing. Additional physiological measurements, such as EMG or heart rate, are necessary to substantiate the claim that freezing is elicited in the paradigm.

To strengthen our claim that freezing is a conditioned response in this task, we have taken three key steps:

(1) We adjusted our freezing detection threshold from 1 cm/s to near 0 cm/s to capture only periods where the animal is virtually motionless on the treadmill. We validated this approach in Figure 2, particularly in the zoomed-in track position trace in Figure 2A, which clearly shows that the identified freezing epochs correspond to no change in track position. All analyses and figures have been updated to reflect this more stringent threshold.

(2) We have added a no-shock control group in the revised manuscript (n = 7; Supplementary Figure 2A-B, H–I) where mice experienced the same protocol, including wearing a tail-coat, but received no shocks. These mice showed no increases in freezing behavior, which further demonstrates that the increased freezing we observe is a result of fear conditioning.

(3) We have added a new supplementary video (Supplementary Video 2) that better illustrates the freezing behavior in our task.

That said, we fully agree with the reviewer that freezing is not the only defensive response observed. Other behaviors—such as hesitation, backward movement, and slowing down—also emerge that are unique to our treadmill-based paradigm. We chose to focus on freezing in this manuscript to align with convention in freely moving fear conditioning studies and to facilitate direct comparisons. We agree that additional physiological measurements (e.g., EMG or heart rate) would provide further validation and could help distinguish between different forms of defensive responses. We view this as an important future direction and plan to incorporate such measures in upcoming studies. We highlight this in the results section (lines 175-179, 262-268) and in the discussion (lines 739-750).

Analysis of Extinction: Extinction dynamics are only analyzed through between-group comparisons within each Recall day, without addressing within-group changes in behavior across days. Statistical comparisons within groups would provide a more robust demonstration of extinction processes.

This is an important distinction and we have now added figures (Supplementary Figures 2H-I, 5C-D, 6C-D) showing within-VR behavior across Recall days, along with statistical comparisons and a description of the extinction process based on these results.

Low Sample Sizes: Paradigm 1 includes conditions with very low sample sizes (N=1-3), limiting the reliability of statistical comparisons regarding the effects of shock number and intensity.

Increasing sample sizes or excluding data from mice that do not match the conditions used in Paradigms 2 and 3 would improve the rigor of the analysis.

While we included all conditions in Figure 2 for completeness, we have separated these conditions in Supplementary Figure 2 to ensure clarity. This allows researchers interested in this paradigm to see the approximate range of conditioned responses observed across different parameters. When comparing Paradigm 1 with Paradigms 2 and 3, we have only used data from 1mA, 6 shocks condition.

Potential Confound of Water Reward: The authors critique the use of reward in conjunction with fear conditioning in prior studies but do not fully address the potential confound introduced by using water reward during the training phase in their own paradigm.

We agree this is a point that needs discussion. We have now noted the limitation of using water rewards during training in the discussion section, particularly its effect on the animal’s motivation in the long term and on place cell activity (lines 814-820).

Recommendations for the authors

Reviewer #1 (Recommendations for the authors):

I suggest changing "3 paradigms" to "3 versions of a CFC paradigm," as the paradigm is fundamentally the same, but parameters were adjusted towards finding an optimal protocol.

We have changed this phrasing where applicable.

Figure S2: There appear to be different sets of shock parameters for different mice, most with an n of 1 or 2. This is not reliable for making a decision for optimal shock parameters and should not be discussed in that way until a full-powered comparison is completed. Also, the N adds up to 19, yet only 18 are described as being included in the study.

We thank the reviewer for this important point. We agree that the current study is not powered to definitively identify optimal parameter settings. We have been careful not to interpret it in that way in the text. Rather, we adopted a commonly used starting point from the freely moving literature—1 mA with six shocks—as our initial condition (lines 196-199). To provide context for others interested in pursuing this work, we have presented a range of conditioned responses from different parameter combinations to illustrate potential variability. In most cases, these data are intended for illustrative purposes only and are not meant to support firm conclusions. We agree that a systematic and fully powered investigation of each parameter would be highly valuable, and we plan to pursue this in future work (and hope other labs contribute to this goal, too), much like the iterative optimizations performed in freely moving paradigms over time.

We thank the reviewer for catching the sample size discrepancy and have now corrected it.

The number of animals for the no-shock condition should be included.

Thank you. We have now included this.

A possible explanation for the lower fear and poorer discrimination in versions 2 and 3 could be that 10 min pre-exposure to the CFC context on day -1 led to latent inhibition. Shorter (or eliminated) pre-exposure may improve outcomes.

We agree that the exposure time is a parameter that we should explore. We have highlighted this in the discussion (lines 729-736) as a parameter that is worth testing in the future.

For analysis of extinction, it is best to establish this within condition - is freezing to the CFC context significantly reduced compared with initial recall and similar to pre-training freezing? By using discrimination as your index of extinction, increases in control context freezing/inactivity can eliminate context discrimination without the conditioned response of freezing actually undergoing extinction.

This is a good point, and we have now included analysis and conclusions based on a within-VR comparison for the analysis of fear extinction (Supplementary Figures 2H-I, 5C-D, 6C-D).

Reviewer #3 (Recommendations for the authors):

Clarification of Treadmill Shape: The manuscript describes the treadmill as "spherical" throughout. However, based on representative images and videos, the treadmill appears cylindrical. This discrepancy should be clarified to ensure consistency between the text and visuals.

The reviewer is correct that the treadmill is cylindrical, and this was an error on our part. We have corrected it throughout.

Figure and Legend Labeling: To improve clarity, all figures and their legends should be explicitly labeled with the corresponding paradigm (1, 2, or 3) to facilitate interpretation.

We have now added a label on all figures that clarifies which Paradigm the figures are referring to. We have also explicitly added this to the figure legends.

Objective Language: Subjective language, such as "since we wanted animals to" (Line 850), should be revised to reflect an objective tone (e.g., "to allow animals to"). Similarly, phrases like "We believe" (Line 896) should be avoided to maintain an unbiased presentation.

We have removed subjective language from our text.

Placement of Future Directions: Speculations on future experimental plans, such as the use of sex as a biological variable (Lines 895-903), should be included in the Discussion section rather than the Methods. Additionally, remarks about the responsiveness of female mice to tail shocks should be moved to the main text for proper contextualization.

We have moved these lines as suggested by the reviewer.

Version 2

Abstract

Abstract

Contextual fear conditioning is a classical laboratory task that tests associative memory formation and recall. Techniques such as multi-photon microscopy and holographic stimulation offer tremendous opportunities to understand the neural underpinnings of these memories. However, these techniques generally require animals to be head-fixed. There are few paradigms that test contextual fear conditioning in head-fixed mice, and none where the behavioral outcome following fear conditioning is freezing, the most common measure of fear in freely moving animals. To address this gap, we developed a contextual fear conditioning paradigm in head-fixed mice using virtual reality (VR) environments. We designed an apparatus to deliver tail shocks (unconditioned stimulus, US) while mice navigated a VR environment (conditioned stimulus, CS). The acquisition of contextual fear was tested when the mice were reintroduced to the shock-paired VR environment the following day. We tested three different versions of this paradigm and, in all of them, observed an increased conditioned fear response characterized by increased freezing behavior. This was especially prominent during the first trial in the shock-paired VR environment, compared to a neutral environment where the mice received no shocks. Our results demonstrate that head-fixed mice can be fear conditioned in VR, discriminate between a feared and neutral VR context, and display freezing as a conditioned response, similar to freely behaving animals. Furthermore, using a two-photon microscope, we imaged from large populations of hippocampal CA1 neurons before, during, and following contextual fear conditioning. Our findings reconfirmed those from the literature on freely moving animals, showing that CA1 place cells undergo remapping and show narrower place fields following fear conditioning. Our approach offers new opportunities to study the neural mechanisms underlying the formation, recall, and extinction of contextual fear memories. As the head-fixed preparation is compatible with multi-photon microscopy and holographic stimulation, it enables long-term tracking and manipulation of cells throughout distinct memory stages and provides subcellular resolution for investigating axonal, dendritic, and synaptic dynamics in real-time.

Introduction

Large-scale multi-photon imaging and holographic stimulation offers the ability to accurately track and record from the same neuronal populations over extended periods at high spatiotemporal resolution, record from axons, dendrites, and dendritic spines, and stimulate neurons with single-cell precision (Bando et al., 2021; Chaigneau et al., 2016; Dombeck et al., 2010; Dong et al., 2021; Emiliani et al., 2015; Fan et al., 2023; Hattori & Komiyama, 2022; Heer & Sheffield, 2024; Hernandez et al., 2016; Krishnan et al., 2022; Russell et al., 2022; Sheffield & Dombeck, 2015). These techniques facilitate the study of neural circuit dynamics, synaptic and plasticity mechanisms and enable the real-time manipulation of neural activity, all of which are crucial for understanding the mechanisms underlying memory formation and recall in the brain. However, head-fixation of animals, which is generally required to utilize these techniques, imposes constraints on the range of testable behaviors compared to freely moving conditions (Guo et al., 2014; Juczewski et al., 2020). This limitation is particularly evident in the Pavlovian contextual fear conditioning (CFC) task, a laboratory method for testing associative learning, fear memory formation, and recall (M. E. Bouton, 1993; Michael S. Fanselow, 1980, 1990; J. J. Kim & Fanselow, 1992; LeDoux, 2000; Maren, 2001; Maren et al., 2013).

In the traditional CFC paradigm for freely moving animals, an animal is placed in an environment (conditioned stimulus, CS) that is paired with an aversive stimulus, such as a mild foot shock (unconditioned stimulus, US). Animals are then removed from the environment. When reintroduced to the environment, animals exhibit a species-specific conditioned response, such as fearful freezing behavior in rodents, if they successfully associate the shock-paired environment with the aversive stimulus. The CFC paradigm is one of the most basic conditioning procedures for freely moving animals— it’s simple and robust (Michael S. Fanselow, 1980, 1990). Despite its longstanding use in studies involving freely behaving animals to explore questions on learning and memory (Anagnostaras et al., 2001; M. E. Bouton & King, 1983; Mark E. Bouton & Bolles, 1980; Debiec et al., 2002; M. S. Fanselow & Bolles, 1979; Fendt & Fanselow, 1999; Gewirtz et al., 2000; Holland & Bouton, 1999; Jeansok J. Kim & Jung, 2006; Maren et al., 2013; Stephen Maren & Holt, 2000; Pezze & Feldon, 2004), it has been noticeably absent in studies involving head-fixed animals, with some exceptions.

Two prior studies (Lovett-Barron et al., 2014; Rajasethupathy et al., 2015) reported a version of CFC in head-fixed mice with aversive air puffs as the US. These studies reported the conditioned response animals displayed as lick suppression, not freezing. Although lick suppression after fear conditioning has also been observed in freely moving animals (Leaf & Leaf, 1966; Leaf & Muller, 1965), it has been shown to result from increased freezing (Mark E. Bouton & Bolles, 1980). Therefore, relying solely on lick suppression in the head-fixed preparation makes it more challenging to compare with the extensive previous work in freely moving animals, where freezing is measured as the predominant conditioned fear response. Furthermore, the aforementioned head-fixed CFC studies also had the limitation of simultaneously having both a water reward and an aversive stimulus within the context, potentially confounding the results and complicating the interpretation. Hence, we aimed to develop a paradigm where freezing is the conditioned response and without the presence of a reward within the conditioned context, making it directly comparable to freely moving animals.

Here, we describe three different versions of a contextual fear-conditioning paradigm for head-fixed mice that resulted in freezing as the conditioned response. The context or CS was a virtual reality (VR)-based environment, which head-fixed mice explored by running on a cylindrical treadmill. In a previous study (Ratigan et al., 2023), we demonstrated the feasibility of this approach by combining it with imaging of thalamic axons innervating the hippocampus using a two-photon microscope. The present manuscript provides a comprehensive description and fundamentals of the behavioral framework, to enable the research community to reproduce and adapt our behavior paradigm. Our main aim is to show that VR-based CFC can elicit freezing responses similar to those seen in freely moving animals, while also examining how key parameters affect this response. Though we explored some important variables, this paper does not attempt to comprehensively investigate all parameters necessary for successful VR-based CFC. Additionally, we performed two-photon imaging of a large population of hippocampal CA1 cells during fear conditioning and tracked the same cells during memory recall across days. We analyzed the impact of our paradigm on place cells in the hippocampus and compared these findings to results from CFC in freely moving animals.

Results

A paradigm for contextual fear conditioning in head-fixed mice exploring virtual reality environments

To establish the contextual fear conditioning protocol (Figure 1A-C), head-fixed mice were water-restricted and trained to run on a treadmill and navigate VR environments for water rewards, as we have demonstrated previously (Dong et al., 2021; Krishnan et al., 2022; Krishnan & Sheffield, 2023; Ratigan et al., 2023). Forward movement on the treadmill translates to forward movement in VR, allowing mice to navigate the VR environment. The environments used were 2m-long linear tracks rich in visual cues (Figure 1). Reaching the track’s end triggered a water reward and defined a completed trial/lap. Mice were then virtually teleported back to the track’s start for the next lap. Unlike freely moving mice that tend to explore new environments, head-fixed mice placed on a treadmill do not automatically start exploring VR environments. However, they can be trained through reinforcement learning using water restriction and water rewards (Dombeck et al., 2010; Dong et al., 2021; Krishnan et al., 2022; Ratigan et al., 2023; Sheffield & Dombeck, 2015).

A contextual fear conditioning paradigm for head-fixed mice navigating virtual reality environments.

(A) Experimental setup created with BioRender.com/sr8vpch. Mice were head-restrained with their feet resting on a cylindrical treadmill. Five large monitors surrounded the mice that displayed virtual reality (VR) environments. Movement on the treadmill advanced the VR display, allowing for context exploration. (B) Mice were water-restricted and trained to run laps in the VR for water rewards. VR environments were 2m-long linear tracks. Mice were trained to achieve >3 laps per minute, which took ∼10 to 14 days (Stage 1). Once well-trained, a “tail-coat” was added to their tails (Stage 2), followed by the removal of the water reward the next day (Stage 3). (C) For Paradigm 1, mice underwent the fear conditioning protocol the following day after the water reward was removed. On the first experiment day (Day 0), mice spent 10 minutes in the training VR (Familiar VR) and then another 10 minutes in a new VR (CFC VR). After the initial exploration, mice received mild electric shocks on the tail (4-12 shocks, 0.5-1.2mA in amplitude, 1s long). The next day (Day 1), mice were tested for memory recall by placing them in the Familiar VR and the CFC VR for 5 minutes each in a counterbalanced manner. (D) Schematic of the tail-coat used for delivering mild electric shocks to the mouse’s tail. (Top) View from the top (Bottom) side view. Dimensions are provided for a typical 12-week-old male mouse weighing ∼30g before water restriction (see Methods for more details).

Training to navigate VR environments on a treadmill (Stage 1), using water rewards, was achieved in approximately 10-14 days, after which well-trained mice could run several laps per minute. We set a criterion of >3 laps per minute (i.e., a lap velocity of 0.1 - 0.2 m/s) for mice to move on to the next stage (Stage 2). Once the mice reached this stage (Stage 2), the following day, we added a coat to their tails (Figure 1B). We developed a lightweight, wearable conductive apparatus for the mouse’s tail to ensure minimal discomfort for the mouse. We named this apparatus the “tail-coat” (Figure 1D). This tail-coat enabled us to administer mild electric shocks to the mouse’s tail (see Methods).

When the tail-coat was added, some mice reduced their running speed or refused to move. Some recovered the next day, but if their running speed did not reach at least three laps per minute (∼20% of mice failed to meet this criterion), they were not advanced to the next stage. Mice that maintained this running speed advanced to Stage 3. Data from mice that advanced with minimal changes to their lap running behavior after adding the tail-coat are shown in Supplementary Figure 1A.

The following day (Stage 3), mice were placed in the same VR environment with the tail-coat, but the water reward was removed (Figure 1B). Mice that continued to run at least three laps per minute with the tail-coat and without water reward advanced to the experiment stage (Stage 4). Around 40% of the mice dropped out at this stage. High levels of lap running during training were crucial to ensure mice were undeterred by both the tail-coat and the reward removal (Krishnan et al., 2022; Krishnan & Sheffield, 2023). Our criteria for consistent running behavior with minimal pauses throughout these three stages of training ensured that any baseline freezing behavior was minimal, similar to freely moving animals, and allowed us to attribute freezing after CFC to fear conditioning. In summary, our criteria for selecting mice for the experimental stage of contextual fear conditioning required them to run consistently (>3 laps per minute) with 1) rewards, then 2) with a tail-coat, and then 3) without rewards plus the tail-coat. This threshold resulted in about 40% of the mice proceeding to the experimental stage. Once mice qualified for the experimental stage, they were no longer excluded from analysis due to their behavior. We tested three versions of the head-fixed CFC paradigm. In Paradigm 1, the neutral environment was the training VR environment, and a new VR environment was paired with shocks. In Paradigms 2 and 3, both the neutral and shock-paired environments were new and distinct from the training environment to better control for environmental familiarity. In Paradigms 1 and 2, the tail-coat was removed on the days following CFC, whereas in Paradigm 3, it remained on the mouse on all post-CFC days. With these paradigms, we varied some parameters that may influence fear memory recall, contextual discrimination, fear extinction, and their associated behaviors in head-fixed VR-based CFC paradigms. We detail these variations in the following sections, starting with Paradigm 1.

When tested for contextual fear memory recall, mice displayed increased freezing behavior in the shocked-paired VR environment.

On Day 0 of the experimental stage in Paradigm 1 (CFC day), mice (n = 27, 25 males and 2 females) were exposed to the same VR environment we used for training, referred to as the Familiar VR from here onwards (Figure 1C), for 10 minutes. We then switched them to a new VR environment (CFC VR). Following baseline exploration of the CFC VR for 10 minutes, we repeatedly applied mild-electric shocks (4-12 shocks, 0.5-1mA in amplitude, 1s long) to the mouse’s tail. We call the period when shocks were delivered “during CFC”. Shocks were pseudo-randomly administered with respect to the animal’s position in the CFC VR, with an inter-shock interval (ISI) of 1 minute. To ensure that the shocks were associated with the context and not a specific position in the environment, no visual cues were explicitly paired with the shock. We quantified the animal’s response to the shocks and observed an immediate increase in running speed following each shock (Supplementary Figure 1B). This was consistent across all mice and shocks and was a reliable indicator that the apparatus was working correctly. Sixty seconds after the last shock was administered, the VR screens were switched off, the tail-coat was removed, and mice were returned to their home cages. The next day (Fear memory recall, Day 1), we tested for recall of the learned association between the VR environment and the fearful stimulus. Mice were exposed to both VRs (familiar VR and CFC VR) for five minutes each in a counterbalanced manner. Water reward remained absent on recall days; however, in this paradigm, the tail-coat was no longer present.

Before fear conditioning, mice exhibited similar behavior in both the Familiar and CFC VRs. We evaluated the time taken to complete a lap and the freezing behavior (defined as time points when the animal’s instantaneous velocity dropped was 0 cm/s). Neither metric revealed a significant difference between the two VRs before shock administration (Figure 2A, Supplementary Figure 1D-E). However, after CFC, when the animals were reintroduced to the VRs on recall Day 1, we observed a significant increase in freezing and time taken to complete a lap in the CFC VR compared to the Familiar VR (Figure 2A-D). As soon as mice transitioned to the CFC VR (either from dark or after the Familiar VR) on the recall day, we observed behaviors such as slowing down, freezing, moving backward, and hesitation to move forward (Supplementary Videos 1-3). This response was markedly different from their behavior in the Familiar VR. This difference in behavior can be observed clearly in most mice on the very first lap (Figure 2A-B). Since we removed the reward, we didn’t find licking behavior to be indicative of recall behavior (Supplementary Figure 1C). However, some mice displayed licking behavior in the Familiar VR on Day 0 before fear conditioning, which was absent on Day 1. This is likely because mice had previously received a water reward in the Familiar VR. We will address this caveat in Paradigms 2 and 3.

Paradigm 1: Head-fixed mice show increased freezing behavior following contextual fear conditioning in the CFC VR.

(A) A single example mouse’s lap running behavior on experiment days, Day 0 and Day 1. Behavior is shown for approximately 3 minutes in all sessions except during CFC, which is shown for the 6 minutes that the session lasted. Frames where freezing was detected (instantaneous velocity 0cm/s) are marked with black dots. A zoomed-in portion on the right in CFC VR highlights these freezing epochs as periods of minimal movement on the treadmill (i.e. instantaneous velocity 0cm/s), which are less visible in the full-scale view. The traces on the right show that this mouse increased freezing, decreased velocity, and moved backward (shown in gray) in the CFC VR (red traces) but not in the Familiar VR (blue traces) on Recall Day. This mouse received six shocks at 1mA intensity at 60 s inter-stimulus interval (ISI). (B) First two minutes of recall behavior in more mice (n = 5) in Familiar VR versus CFC VR. (C-D) (Left Average freezing percentage on recall day in the very first lap (C) and all laps (D) during the 5 minutes that mice explored the Familiar (blue) and CFC (red) VR. Freezing (%) was calculated as the number of frames where freezing was detected in a lap divided by the total number of frames in each lap. (Right) Delta calculated as the difference in the amount of freezing in the CFC VR compared to the Familiar VR before CFC (Day 0) and after CFC (Day 1). The dashed line represents 0. (E-F) Same as C-D but for the time taken to complete the first lap (E) and all laps (F). Mice displayed an increase in freezing and in time taken to complete a lap in the first lap and, on average, in the CFC VR. In C-F, circles and pluses represent individual mice (n = 27, 25 male and two female mice). In C-F, data was pooled from mice receiving different numbers of shocks (4, 6, 12) at varying intensities (0.5 mA, 0.6mA, 0.8 mA, 1 mA, and 1.2 mA), which is separately displayed in Supplementary Figure 2. Lines join data from the same mouse. P-values were calculated using a paired t-test.

Running and licking behaviors in Paradigm 1.

(A) Box plot of average time taken to complete a lap before and after the addition of a tail-coat recorded in a randomly chosen subset of mice (circles, n=8). (B) Instantaneous velocity increased immediately after shock onset. This increased running speed also served as a reliable measure to confirm that mice received the tail shock. The grey line indicates the average velocity calculated by shuffling the instantaneous velocity 100 times. (C) Licking behavior doesn’t stop immediately in the familiar VR without water, as mice previously received water rewards in the familiar VR. However, it was low in the other sessions and across environments. There was no significant difference between the Familiar and CFC VR on Day 1. Licking behavior was collected only in a subset of mice (circle, n=5). (D-E) The amount of freezing (D) and time taken to complete a lap (E) were not significantly different between the Familiar and CFC VR before fear conditioning (n=25). While not significantly different, there was a trend towards longer time taken in the first lap of the CFC VR (E), which we have observed before when mice enter novel environments from familiar ones (Dong et al., 2021). The significantly lesser freezing during this first lap (D) suggests that animals were exploring slowly rather than remaining immobile. The boxplots (in A, C-E) range from the first quartile (25th percentile) to the third quartile (75th percentile), and the box shows the interquartile range (IQR). The line across the box represents the median (50th percentile). The whiskers extend to 1.5*IQR on either side of the box, and anything above this range is defined as an outlier. P-values were calculated using a paired t-test.

Change in freezing with shock amplitude, number of shocks, and across recall days in Paradigm 1.

(A-B) The left panel shows freezing in the first lap (A) and across all laps (B). The right panel illustrates delta freezing, calculated as the difference in freezing between the CFC VR and the Familiar VR. The colors represent the shock parameters used. Number of animals used is indicated in the figure legend. The dashed line represents the average freezing percentage in a control paradigm, where mice experienced the same conditions without shocks (n = 7). In most cases, we observed an increase in freezing in the CFC VR compared to the Familiar VR (delta>0). (C-D) Freezing in the first lap (C) and across all laps (D) by shock amplitude. In all cases, mice received six shocks that were 1 minute apart. Greater fear discrimination could be seen at higher shock amplitudes. (E) Median freezing in the CFC VR by shock amplitude for first lab freezing (black) and average freezing across laps (gray). (F-G) A subset of animals underwent multiple days of recall tests (n = 25 Day 0, n = 27 Recall Day 1, n = 13 Recall Day 2, n = 8 Recall Day 3, n = 7 Recall Day 4). VR environments were presented in a counterbalanced manner for 5 minutes each day. The increase in freezing in the CFC VR was highest on the first day, both in the first lap (F) and across all laps (G). P-values were calculated using a paired t-test. (H-I) Within-VR comparison of the same dataset. First lap freezing (H) and average freezing across all laps (I), separated by Familiar VR (left panels) and CFC VR (right panels), demonstrate fear extinction across days. The black line represents average freezing in control mice that received no shocks. P-values indicate significant differences between experimental and no-shock control groups (t-test). Fear extinction is quantified by the LME model in Tables 1 and 2, which shows the statistical comparison of freezing in recall days versus Day 0 (baseline) within each VR.

To quantify the distinct conditioned responses observed (Figure 2A-B, Supplementary Videos 1-3), we used two metrics: freezing behavior, defined as periods when the animal’s instantaneous velocity was 0 cm/s, and time taken to complete a lap, which captured the avoidance-like behaviors (such as slowing down, moving backward, and hesitation to move forward) as well as freezing. When we quantified the freezing behavior in mice, we found it to be, on average, high throughout the five minutes that mice spent in the CFC VR on recall Day 1 (Figure 2C-D, average freezing (%), mean ± 95% Confidence Intervals (CI), Familiar VR: 17.52 ± 3.25, CFC VR: 23.54 ± 3.26, p=0.001, Paired t-test, n = 27). We also observed a significant increase in freezing during the first lap in the CFC VR compared to the familiar VR (Figure 2B-C, average first lap freezing (%), Familiar VR: 15.94 ± 4.03, CFC VR: 25.11 ± 3.41, p<0.001, Paired t-test). Mice also took longer to complete a lap in the CFC VR compared to the Familiar VR, both in the first lap (average time taken to complete first lap (s), Familiar VR: 14.81 ± 3.74, CFC VR: 41.16 ± 9.33, p<0.001, Paired t-test) and on average (Figure 2E-F, the average time taken to complete all laps (s), Familiar VR: 16.17 ± 3.05, CFC VR: 38.82 ± 11.06, p<0.001, Paired t-test). The increase in freezing and time taken to complete a lap in the CFC VR compared to the Familiar VR was significantly higher on Day 1 than Day 0, indicating that it results from fear conditioning (Figure 2C-F, p<0.05, Paired t-test).

Based on studies in freely moving animals (Michael S. Fanselow, 1980, 1990; Ramanathan et al., 2018; Russo & Parsons, 2021; Trott et al., 2022), we settled on an intensity of 1 mA and six shocks as a reasonable protocol, which we tested on a sizeable cohort of mice (n = 12). We found that freezing at this intensity and shock number was higher on average in the CFC VR; however, mice showed variability in their responses (Supplementary Figure 2A-B). The majority of mice (7/12) displayed appropriate memory recall (Supplementary Figure 2A-B), indicated by more freezing in the CFC VR compared to the Familiar VR (delta > 0), while 3/12 mice froze more in the Familiar VR (delta < 0), and 2/12 showed equal freezing levels in both VRs (delta ≈ 0). Such variability has been regularly documented in freely moving animals and may not be unique to the VR setup used here (Chu et al., 2024; Dos Santos Corrêa et al., 2019; Navarro-Sánchez et al., 2024; Poulos et al., 2016; Russo & Parsons, 2021; Totty et al., 2021).

Next, in a small group of mice (n = 15), we investigated whether varying the number of shocks or the shock amplitude would affect the conditioned fear response (Supplementary Figure 2A-B). Though our parameter exploration was not exhaustive, we assessed shock amplitude more extensively than the number of shocks. (Supplementary Figure 2C-E). Overall, most shock amplitudes and numbers of shocks increased freezing behavior in the CFC VR on recall day (Supplementary Figure 2A-B). The observed variability was within the same range as that of the larger cohort of mice we tested using the 1 mA, 6-shock protocol (Supplementary Figure 2A-B). Notably, increasing the shock amplitude from 0.6 mA to 1 mA, while keeping the number of shocks consistent at 6 shocks, significantly enhanced both fear discrimination and freezing behavior (Supplementary Figure 2C-E). A more detailed analysis with a larger sample size may be necessary to understand further how variations in shock parameters affect the conditioned response, as has been done in studies with freely moving animals (Baldi et al., 2004; Cordero et al., 1998; Daviu et al., 2024; Poulos et al., 2016; Quinn et al., 2008).

Finally, we examined the evolution of this conditioned response across days (Supplementary Figure 2F-I). We placed the animals in both the Familiar and the CFC VR for five minutes over four days (Recall Days 1-4) in a counterbalanced manner and tested their recall behavior. We quantified both the degree of fear discrimination across days (Supplementary Figure 2 F-G) and the extinction of the fear response within the CFC VR compared to baseline (Supplementary Figure 2 H-I).

Fear discrimination was most prominent on Recall Day 1. Mice displayed the most substantial increase in freezing in the CFC VR compared to the Familiar VR on Recall Day 1 (Supplementary Figure 2F-G). Freezing on the first lap in the CFC VR was also most pronounced on Recall Day 1. Both average and first lap freezing in the CFC VR decreased on Recall Day 2 compared to Recall Day 1, becoming similar to the Familiar VR. While first lap freezing remained similar between the Familiar and the CFC VR on subsequent Recall Days 3 and 4, we observed some variability in the average freezing behavior. Average freezing increased significantly again on Recall Day 3 in the CFC VR but was similar in both VRs on Recall Day 4.

To determine the extinction of the fear response, we compared freezing within each VR to baseline freezing levels on Day 0 (Supplementary Figure 2H-I). We quantified this using a Linear Mixed Effects (LME) Model, treating recall days as fixed effects and mouse as a random effect, with baseline freezing on Day 0 as the reference (Tables 1-2). In the CFC VR, both first-lap and average freezing were most prominent on Recall Day 1 and returned to baseline levels by Recall Day 2. While we observed a small increase in average freezing on Recall Day 3, it remained lower than Recall Day 1 levels and returned to baseline by Recall Day 4. This pattern suggests that the pronounced first-lap freezing observed on Recall Day 1 extinguishes quickly, while average freezing, though more variable, also extinguishes by Recall Day 4.

In the Familiar VR, freezing behavior—both first lap and average—did not differ significantly from baseline on most days. However, we observed an unexpected increase in average freezing compared to baseline on Recall Day 4. This increase could reflect either fear generalization or decrease in motivation - a response to the prolonged absence of reward in a previously rewarded environment. To distinguish between these possibilities, we examined freezing behavior in control mice that underwent the same paradigm without shocks (no-shock controls, n = 7, black lines in Supplementary Figure 2H-I). Average freezing in the Familiar VR exceeded that of no-shock controls on both Recall Days 1 and 4. Since no-shock controls explored the environments under no-reward conditions, this suggests that the increase in freezing in the Familiar VR stems from fear conditioning and may indicate generalization rather than motivational changes. This aligns with findings from freely moving animals, where specific fear memories have been shown to generalize over time with repeated exposures to both neutral and conditioned contexts (Biedenkapp & Rudy, 2007; Wiltgen & Silva, 2007).

In summary, these results demonstrate that head-fixed mice can acquire fear associations in VR environments. Most mice displayed freezing as a conditioned response specifically to the shock-paired VR and showed robust fear discrimination—behavior that mirrors that of freely moving animals after fear conditioning. By the fourth recall day, freezing had largely extinguished. Notably, beyond freezing, other avoidance-like defensive behaviors were prominent in the shock-paired VR, demonstrating the range of behavioral responses our paradigm can capture.

A second paradigm where head-fixed mice discriminate between two novel VRs

As previously mentioned, the incomplete extinction of licking behavior in the familiar VR before fear conditioning or the gradual increases in freezing behavior in the familiar VR during recall tests may indicate that mice still associate the VR with water rewards, which could be a caveat. Furthermore, in traditional CFC paradigms with freely moving mice, fear behavior has been assessed mainly by comparing two novel contexts, one associated with shocks and the other acting as a control. To address this, we tested another paradigm (Paradigm 2), using a novel VR instead of the familiar VR to act as the control (Figure 3A). One novel VR was assigned as the Control VR, and a second novel VR was assigned as the CFC VR, which would later become paired with the shocks. In this paradigm, we also added an extra habituation day to increase the animal’s pre-exposure time to the novel VRs prior to conditioning (Day -1). Pre-exposure to the context before fear conditioning can enhance contextual fear, as seen in freely moving animals (Biedenkapp & Rudy, 2007; Michael S. Fanselow, 1980, 1990; Maren et al., 2013; Rudy et al., 2002; Wiltgen & Silva, 2007).

Using a novel VR as the neutral environment instead of a familiar VR in a second paradigm results in modest increases in freezing.

(A) The training paradigm for Paradigm 2 is similar to Paradigm 1. On experiment day, mice were introduced to two novel VRs, one of which would be associated with the shock (CFC VR) and the other wouldn’t (Control VR). In this paradigm, there was an added habituation day, where mice were exposed to the two VRs for ten minutes. The next day, mice ran in the two VRs again before receiving mild electric shocks in the CFC VR. The recall test occurred the next day. In this cohort, all mice received six shocks at 1 mA intensity. (B-D) These panels show the lap running behavior of a single mouse on experiment days -1, 0, and 1. Behavior is displayed for about 3 minutes in all sessions except during CFC. Frames with freezing detected by a threshold (instantaneous velocity 0cm/s) are marked by black dots. In (D), this mouse shows an increase in freezing, a decrease in velocity, and some backward movement when in the CFC VR (red traces). (E) Comparison between freezing in the first lap (left) and all laps (right) in Paradigm 1 (blue and red) versus Paradigm 2 (green and red). In Paradigm 1, only animals that received six shocks at 1 mA intensity are included. n = 12 mice were used in both paradigms (10 male and two female mice). The scale bar for panels B-D is indicated in B. P-values were calculated using a paired t-test.

The training was done similarly to the previous paradigm. After mice showed consistent running behavior in the Familiar VR with the tail-coat and without reward, they were advanced to the experimental stage. On Day -1, mice were exposed to the two novel VRs for ten minutes each. On Day 0, they were re-exposed to the same two VRs for ten minutes before receiving six one-second-long mild electric shocks (at 1 mA and 1 minute ISI) in one of them - the CFC VR. On Day 1, mice were exposed to the two VRs for five minutes to evaluate memory recall (Figure 3A-B).

Compared to Paradigm 1, in the novel VRs, mice showed no differences in licking behavior across all experimental sessions (Supplementary Figure 3A). There was also no significant difference in freezing behavior and time taken to complete a lap between the two VRs on Day -1 and Day 0 before fear conditioning (Supplementary Figure 3B-C).

Head-fixed mice show heightened fear discrimination during the first lap when discriminating between two novel VRs.

We compared the freezing behavior between Paradigm 1 and Paradigm 2 on Recall Day 1 to the same intensity and number of shocks (1 mA, 1 s long, 60 s ISI, six shocks, n=12 mice each paradigm). While first-lap freezing in the CFC VR was markedly higher in Paradigm 2, significant differences between the Control VR and CFC VR were observed only for first-lap freezing, not for average freezing or lap completion time across all laps (Figure 3B-D, 4A-E). The effect sizes for the difference in first-lap freezing between the two VRs were nearly identical between the paradigms (Cohen’s d: First-lap freezing: Paradigm 1: 1.038, Paradigm 2: 1.040), but the effect size for average freezing was notably smaller in Paradigm 2 compared to Paradigm 1 (Cohen’s d: Average Freezing: Paradigm 1: 0.797, Paradigm 2: 0.093).

When we quantified freezing by lap on Recall Day 1 in Paradigm 2, we found that after the first lap, mice showed similar freezing in both the Control VR and the CFC VR across all subsequent laps (Supplementary Figure 4). In contrast, in Paradigm 1, mice consistently froze more in the CFC VR compared to the Familiar VR throughout Recall Day 1 (Supplementary Figure 4).

Indeed, when we quantified freezing behavior in the Control VR and CFC VR in Paradigm 2, we found that the most significant increase in freezing and other avoidance-like defensive behaviors in the CFC VR was observed in the first lap (freezing in first lap %: Control VR: 11.04 ± 1.8 CFC VR: 16.05 ± 3.94, p = 0.016, time taken to complete first lap (s): Control VR: 14.27 ± 2.85, CFC VR: 25.68 ± 5.55, p = 0.001, Paired t-test, Figure 3B-E, Figure 4A-E). Neither average freezing across all laps (Figure 4C, average freezing in all laps %: Control VR: 20.54 ± 4.13, CFC VR: 21.15 ± 4.21, p=0.738, Paired t-test) nor time taken to complete all laps (Figure 4E, the average time taken to complete all laps: Control VR: 21.73 ± 8.11, CFC VR: 24.27 ± 6.36, p=0.558, Paired t-test) showed significant differences between the two VRs.

In Paradigm 2, the most significant freezing and reduced speed occurred in the first lap of the CFC VR compared to the Control VR.

(A) First two minutes of recall behavior in five mice in Control VR (green) versus CFC VR (red). (B-C) The left panels show the amount of freezing on the recall day in (B) the very first lap and (C) all laps during the 5 minutes when the mice explored the Control (green) and CFC (red) VR. Freezing (%) is calculated as the number of frames where freezing was detected in a lap divided by the total number of frames in each lap. On the right, the delta is calculated as the difference in the amount of freezing in the CFC VR compared to the Control VR before (Day 0) and after CFC (Day 1). The dashed line represents 0. (D-E) Same as B-C but for the time taken to complete the first lap (D) and all laps (E). In B-E, circles and pluses represent individual mice (n = 12). Lines join data from the same mouse. P-values were calculated using a paired t-test.

Licking and running behavior before fear conditioning, in Paradigm 2.

(A) Licking behavior remains low in all sessions, unlike Paradigm 1. It does not significantly differ between the Control vs CFC VR before and after CFC. (B-C) The amount of freezing and average running speed was not significantly different between the Control and CFC VR before fear conditioning, both on Day -1 (B) and Day 0 (C). P-values were calculated using a paired t-test.

Lap-wise freezing behavior in the three paradigms.

(A-F) Average freezing across mice (A-C) and the number of freezing epochs (D-F, see Methods) in each lap across the two VRs on the first day of recall in (A, D) Paradigm 1, (B, E) Paradigm 2 and (C, F) Paradigm 3. Lines indicate mean, and shading indicates 95% confidence intervals. P-values between (A, D) Familiar VR and CFC VR and (B-C, E-F) Control VR and CFC VR were calculated using a paired t-test. * indicates p-values<0.01.

Mice largely extinguished their fear within the first day of recall in Paradigm 2.

(A-B) First lap freezing (A) and average freezing across all laps (B) across two days of recall in Paradigm 2. P-values were calculated using a paired t-test. Circles indicate data from individual mice (n = 12 for Recall Day 1 and n = 8 for Recall Day 2). P-values were calculated using a paired t-test. (C-D) Within-VR comparison of the same dataset. First lap freezing (C) and average freezing across all laps (D), separated by Familiar VR (left panels) and CFC VR (right panels), demonstrate fear extinction across days. Fear extinction is quantified by the LME model in Tables 3 and 4, which shows the statistical comparison of freezing in recall days versus Day 0 (baseline) within each VR.

When mice were returned to the two VRs the following day for a second day of recall (Recall Day 2), they continued to exhibit modest levels of fear discrimination, as indicated by a small but statistically significant difference in freezing in the first lap and on average between the CFC VR and Control VR (Supplementary Figure 5 A-B). However, when assessing fear extinction by comparing freezing in the CFC VR across recall days to baseline levels (Supplementary Figure 5 C-D, Tables 3-4), we found no significant differences beyond the very first lap on Recall Day 1. This suggests that while mice maintained their ability to discriminate between environments by freezing more in the CFC VR, their overall fear response had extinguished. Thus, in Paradigm 2, fear extinction occurs rapidly within the first recall session after the initial lap.

A shorter ISI and leaving the tail-coat on during fear memory recall improved fear discrimination when using two novel VRs.

Freezing behavior in the CFC VR in Paradigm 2 was weaker than in Paradigm 1, with rapid fear extinction after the first recall lap. In Paradigm 3, two parameters were changed relative to Paradigm 2 to test whether freezing behavior and fear discrimination could be enhanced. First, the interval between consecutive shocks was reduced to see if a shorter ISI could produce a more robust fear response (McNally & Westbrook, 2006; Williams, 1994). Second, the tail-coat was kept on during memory recall, based on evidence that removing contextual cues can reduce conditioned freezing (González et al., 2003). We hypothesized that the tail-coat might act as a contextual cue and that its presence during recall might enhance freezing behavior. To further test these effects, we used a lower shock intensity of 0.6 mA—which we previously showed as producing only modest fear responses in Paradigm 1 (Supplementary Figure 2 C-E)—to determine whether a shorter ISI and the retention of the tail-coat during recall could enhance conditioned fear responses produced by the lower shock intensity.

In Paradigm 3, mice followed the same experimental protocol as Paradigm 2, with two novel VRs: one assigned as the Control VR and the other as the CFC VR. A habituation day in the two new VRs (Day -1) was followed by the CFC day (Day 0), where mice received six shocks (0.6 mA, 20-26 s ISI) in the CFC VR. This was followed by recall days (Days 1-3), where mice explored the two VRs presented in a counter-balanced manner. Unlike the other two paradigms, in Paradigm 3, the tail-coat remained on during the recall days (Figure 5A).

A third paradigm that uses a novel VR as the neutral environment but keeps the tail-coat on during memory recall led to increased freezing in the CFC Environment.

(A) The training and experiment paradigm was similar to Paradigm 2 except that the shocks were administered closer together (20-25 s ISI), and the tail-coat was kept on during recall days. (B) First two minutes of recall behavior in Paradigm 3 in five mice in Control VR (green) versus CFC VR (red). (C-D) Comparison between freezing in the first lap (C) and all laps (D) in Paradigm 1 (blue and red) versus Paradigms 2 and 3 (green and red). n = 12 mice were used in Paradigms 1 and 2 (10 male and two female mice), and n = 20 male mice in Paradigm 3. P-values were calculated using a paired t-test. In Paradigm 3, by keeping the tail-coat on during the recall days, we observed an increase in freezing in the CFC VR in the first lap and across all laps compared to Paradigm 2.

We found that mice (n = 20 male mice) showed better fear discrimination between the Control and CFC VRs in Paradigm 3 compared to Paradigm 2 (Figure 5B-D). Mice not only showed significant freezing in the first lap (freezing in first lap %: Control VR: 13.61 ± 2.24 CFC VR: 19.73 ± 5.17, p = 0.021, Paired t-test) but average freezing across all laps was also higher in the CFC VR compared to the Control VR (Figure 5C-D, Figure 6A-D, average freezing in all laps %: Control VR: 15.28 ± 2.00, CFC VR: 21.92 ± 2.95, p < 0.001, Paired t-test).

Mice exhibited increased freezing and overall reduced speed in the CFC VR in Paradigm 3

(A-B) The right panels show the amount of freezing on the recall day in (A) the very first lap and (B) all laps during the 5 minutes when the mice explored the Control (green) and CFC (red) VR. Freezing (%) is calculated as the number of frames where freezing was detected in a lap divided by the total number of frames in each lap. On the left, the delta is calculated as the difference in the amount of freezing in the CFC VR compared to the Control VR before (Day 0) and after CFC (Day 1). The dashed line represents 0. (C-D) Same as A-B but for time taken to complete the first lap (C) and all laps (D). Circles and pluses represent individual mice (n = 20). Lines join data from the same mouse. P-values were calculated using a paired t-test.

Mice displayed better fear discrimination and a delay in extinction in Paradigm 3 compared to Paradigms 1 and 2

(A-B) Freezing in the first lap (A) and average freezing across all laps (B) in the Control VR vs CFC VR on three consecutive recall days in Paradigm 3 (n = 20). The tail-coat was kept on all recall days. P-values were calculated using a paired t-test. (C-D) Within-VR comparison of the same dataset. First lap freezing (C) and average freezing across all laps (D), separated by Familiar VR (left panels) and CFC VR (right panels), demonstrate fear extinction across days. Fear extinction is quantified by the LME model in Tables 5 and 6, which shows the statistical comparison of freezing in recall days versus Day 0 (baseline) within each VR.

While first-lap freezing showed a weaker effect size in Paradigm 3 compared to the other paradigms, there was a strong effect of average freezing across the session (Cohen’s d: First lap freezing: Paradigm 1: 1.038, Paradigm 2: 1.040, Paradigm 3: 0.719. Average Freezing: Paradigm 1: 0.797, Paradigm 2: 0.093, Paradigm 3: 1.232). On average, we found that Paradigm 1 still produced the highest absolute freezing behavior in the CFC VR (average freezing in all laps %: CFC VR: Paradigm 1: 23.54 ± 3.26, Paradigm 2: 21.51 ± 4.21, Paradigm 3: 21.92 ± 2.95). The difference in freezing behavior between the CFC VR and the neutral VR was comparable between Paradigms 1 and 3 (Δ average freezing % (CFC VR – Familiar/Control VR): Paradigm 1: 6.74, Paradigm 2: 0.61, Paradigm 3: 6.63). Thus, unlike Paradigm 2, where the largest differences in freezing between the CFC VR and the Control VR were restricted to the first lap, the shorter ISIs and the addition of the tail-coat during the recall day in Paradigm 3 improved contextual fear discrimination between the two novel VRs. When tested for memory recall in Paradigm 3, mice froze more in the CFC VR than in the Control VR, and this effect persisted beyond the first lap (Figure 5C-D, Supplementary Figure 4). Mice also maintained their ability to discriminate between the two VRs for the first two recall days, freezing significantly more in the CFC VR than in the Control VR on both Recall Days 1 and 2 (Supplementary Figure 6A-B). However, by Recall Day 3, this discrimination diminished, as freezing levels no longer differed significantly between the two VRs.

Unlike the rapid fear extinction observed in Paradigm 2, mice in Paradigm 3 showed delayed extinction (Table 5-6, Supplementary Figure 6 C-D). Freezing levels in the CFC VR remained elevated above baseline during the first lap and across all laps on Recall Days 1 and 2, with average freezing still elevated above baseline on Recall Day 3. While freezing levels showed a consistent downward trend across days, these results suggest that fear had not completely extinguished by Recall Day 3. As in Paradigm 1, the most pronounced increases in first lap freezing in the CFC VR occurred in early recall days, while elevated average freezing persisted longer. These findings suggest that the shorter ISI and presence of the tail-coat during memory recall may enhance conditioned fear and delay fear extinction in head-fixed mice discriminating between two novel VRs, though the individual effects of these parameters require further investigation.

The tail-coat’s presence during recall days likely contributed to some transfer of fear to the Control VR, evidenced by elevated freezing in the Control VR compared to baseline. Nevertheless, freezing levels in the Control VR remained consistently lower than in the CFC VR across all days (Supplementary Figure 6, Tables 5-6).

In summary, our results show that mice exhibited greater average freezing behavior in the CFC VR in Paradigms 1 and 3 compared to Paradigm 2. In Paradigm 2, the most noticeable increase in freezing occurred during the very first lap in the CFC VR compared to the Control VR. Fear extinction was rapid in Paradigm 2 and was back to baseline levels within the first day of recall. We found that fear discrimination could be improved, and fear extinction could be delayed with a shorter ISI when presenting tail shocks and when the tail-coat was present during memory recall, as shown in Paradigm 3. Thus, we provide three different variations for performing CFC in head-fixed mice that can be used to understand the neural underpinnings of fear memory.

Place cells in the CA1 subregion of the hippocampus displayed remapping and decreased field widths following contextual fear conditioning

The hippocampus plays a critical role in the encoding, consolidation, and recall of memories and is necessary for the expression of contextual fear memory (Anagnostaras et al., 2001; Ji & Maren, 2007; Liu et al., 2012; Maren et al., 2013; Stephen Maren & Holt, 2000; Phillips & LeDoux, 1992; Ramirez et al., 2013; Wiltgen & Silva, 2007). Context is thought to be represented in the hippocampus by place cells that fire at specific spatial locations (place fields) in an environment (O’Keefe & Dostrovsky, 1971). These place cells have been thought to help discriminate between feared and neutral contexts. Previous studies on freely moving animals have shown that fear conditioning can cause place cells to remap (Blair et al., 2023; E. J. Kim et al., 2015; Kinsky et al., 2023; Mamad et al., 2019; Moita et al., 2004; Ormond et al., 2023; Schuette et al., 2020; Wang et al., 2012, 2015; Wu et al., 2017). Remapping is defined by place cells shifting their firing fields, indicative of place cells adapting to incorporate new information into memory (Colgin et al., 2008; Leutgeb et al., 2005; Muller & Kubie, 1987). Additionally, a recent study (Schuette et al., 2020) that performed calcium imaging of hippocampal cells in freely moving mice found that fear conditioning narrows the width of place fields in the feared environment. This suggests that place cells encode the same environment on a finer scale following fear conditioning, potentially enhancing context discrimination and threat avoidance (Schuette et al., 2020).

We replicated these findings in our head-fixed version of CFC in VR. To investigate the effect of fear conditioning on place cell activity in our task, we expressed a calcium indicator, GCaMP6f (Chen et al., 2013), in hippocampal pyramidal cells and, using two-photon microscopy, imaged from the same hippocampal neurons across days during fear conditioning and fear memory recall in Paradigm 1 (Figure 7A-B, n=8 mice). We found that a subset of place cells formed before CFC remained stable, while others remapped or changed their preferred firing locations after CFC (see examples in Figure 7C, 7D-G). This is similar to previously reported observations (Blair et al., 2023; Moita et al., 2004; Wang et al., 2012, 2015).

Fear conditioning results in place cell remapping in both Familiar and CFC VR.

(A) A schematic of the procedure for two-photon imaging of large populations of the same CA1 neurons over multiple days, created with BioRender.com/vhiqcof (B) An example field of view across the two experiment days. The same field of view was aligned and imaged across days. (C) Examples of a few place cells in the CFC VR across sessions on Day 0 (before CFC, during CFC) and Day 1 (Memory Recall). White lines separate laps in each session. Some place cells maintain stable fields across days, while others remap. (D) Place fields defined on Day 0 in Familiar VR before CFC are plotted across Day 1 during fear memory recall (n = 8 mice). (E) Place fields defined on Day 0 in CFC VR before CFC are plotted across the fear conditioning session and on Day 1 during fear memory recall. (F) On the left is a scatter plot of the center of mass of place fields defined in Familiar VR before CFC on Day 0 (x-axis) compared to their center of mass on Day 1 during fear memory recall (y-axis). On the right is a boxplot of correlation coefficients between mean place fields defined in Familiar VR before CFC on Day 0 and memory recall on Day 1 (Day 0: Day 1). The within-session correlation coefficients serve as control and were calculated between mean place fields in the first half and second half of the Familiar VR before the CFC session. (G) A scatter plot of the center of mass of place fields defined in CFC VR before CFC on Day 0 (x-axis) compared to their center of mass on (right) Day 0 during CFC and (middle) Day 1 during memory recall. On the left is a distribution of correlation coefficients between mean place fields defined in Familiar VR before CFC on Day 0 and during CFC (Day 0: Day 0) and memory recall on Day 1 (Day 0: Day 1). The within-session correlation was calculated between mean place fields in the first half and second half of the CFC VR before the fear conditioning session. Asterisk (*) indicates significant P values (KS test, P < 0.01). Our findings show that place fields present in the before CFC sessions in both Familiar and CFC VR showed significant remapping following fear conditioning.

We found remapping to occur in both the Familiar and the CFC VR to similar extents, as indicated by the lower correlation across sessions/days compared to the within-session correlation (Figure 7F-G). This contradicts previous findings by Moita et al. (2004), who reported more stability in the control environment compared to the CFC environment after fear conditioning. This discrepancy may be explained by the different reward protocols used. Moita et al.’s rats foraged continuously for food pellets during the experimental sessions, while our mice received no rewards. We removed water rewards from the previously rewarded Familiar VR, and as we have shown in earlier work, removing rewards from a previously rewarded environment causes the place map to become more unstable and likely to remap across days (Krishnan et al., 2022; Krishnan & Sheffield, 2023). Additionally, Moita et al. measured place cell stability within a single day, with just one hour between conditioning and retrieval sessions, while our recordings were separated by a day, including intervening sleep periods. This longer interval may have contributed to more pronounced remapping compared to the shorter interval used by Moita et al (Cai et al., 2009; Maboudi et al., 2018).

Next, we looked at the place cells in each session. Interestingly, we found an increase in the number of place cells in the CFC VR during memory recall, a finding not previously reported (Figure 8A-B). We also found that place cells had narrower fields in the CFC VR during memory recall (Figure 8C), as previously shown (Schuette et al., 2020). We did not find a significant difference in other place cell parameters we examined, such as lap-by-lap reliability (Figure 8D) and out-of-field firing (Figure 8E). Thus, our observations of remapping and narrower place field widths replicate findings from freely moving animals, validating our paradigm’s use to explore further questions using the head-fixed preparation to study the neural underpinnings of contextual fear memories. We add to this field by showing an increase in the proportion of place cells following CFC, a possible neural substrate of fear memory recall.

The widths of place fields in CA1 narrow during memory recall.

(A) Place fields were defined and sorted by track length in each session, pooled from all mice (n = 8 mice). Each cell’s activity was normalized to its peak, and cells were sorted by their center of mass along the track. (B) The percentage of place cells is calculated as the number of place cells divided by the total number of recorded cells. More place cells were identified by our algorithm in the CFC VR on Day 1. (C-E) Parameters of place fields: (C) Width, (D) Place Field (PF) Reliability, and (E) out/in field firing ratio in both VRs on Day 0 and Day 1. The width of the place fields in the CFC VR significantly decreased on Day 1. No other parameters significantly varied across days. P-values were calculated using a paired t-test.

Discussion

We present three paradigms for performing CFC in head-fixed mice, using a conductive “tail-coat” to deliver aversive shocks as mice navigate VR environments on a treadmill. Our key findings are: 1) Head-fixed mice exhibited freezing behavior in response to an aversive stimulus in VR, consistent with conditioned responses seen in freely moving animals. Freezing was especially pronounced during the first lap in the fear-conditioned environment compared to the neutral environment across all paradigms tested. 2) With mice on a cylindrical treadmill, additional conditioned responses such as hesitation and backward movement were detected. 3) The choice of neutral environment influenced fear discrimination: mice froze more in the fear-conditioned VR when the neutral environment was familiar rather than novel. Freezing in the fear-conditioned VR could be enhanced by shorter ISIs and the presence of the tail-coat during recall when the neutral environment was novel. 4) Following fear conditioning, place cells in the hippocampal CA1 region showed remapping, increased numbers, and narrower fields, paralleling findings in freely moving animals and offering new insights.

Here, we experimented with three paradigms for fear conditioning in mice, with the main difference being the neutral environment where the mice didn’t receive any shock. In Paradigm 1, the neutral VR was familiar, where the mice had previously received rewards, and the CFC VR was novel. In Paradigms 2 and 3, both the neutral and CFC VRs were similarly novel. In Paradigm 3, we decreased the ISI and kept the tail-coat on the animal when testing for memory recall. Thus, we have tested three different versions for performing CFC in head-fixed mice. On average, mice showed more freezing behavior in the CFC VR compared to the neutral VR. A common feature across all paradigms was pronounced freezing during the very first lap in the CFC VR. While mice in Paradigm 2 extinguished fear rapidly after the first lap, freezing behavior persisted up to 3-4 recall days in Paradigms 1 and 3, to varying degrees. We also observed variability in conditioned responses across mice. Some displayed an opposite pattern, freezing more in the neutral VR or equally in both VRs. This variability in behavior, similar to that observed in freely moving animals (Chu et al., 2024; Dos Santos Corrêa et al., 2019; Navarro-Sánchez et al., 2024; Poulos et al., 2016; Russo & Parsons, 2021; Totty et al., 2021), can be insightful when combined with investigations of neural activity, helping to understand the neural dynamics that contribute to an animal’s ability to learn the association between a context and fearful stimuli. This can reveal the mechanisms by which such memories may generalize or become maladaptive, as seen in post-traumatic stress disorders.

We have shown here an example of using Paradigm 1 to study place cells in the hippocampus and their role in fear memory formation and recall. We recorded from a large group of the same hippocampal CA1 cells across days and demonstrated that place cells remap and narrow their place field widths with fear conditioning. This occurs in head-fixed mice navigating VR environments, similar to freely moving animals. Narrow place fields may suggest sharper spatial tuning to incorporate a salient emotional memory and may aid in context discrimination (Schuette et al., 2020). However, how place cells narrow their widths in response to fear conditioning remains an open question. One possible way to answer this is by imaging CA1 place cell dendrites to understand how they integrate inputs before and after fear conditioning. We can explore what dendritic firing and plasticity mechanisms lead to narrow place field widths. We can then perturb these firing patterns to attribute causality to the narrowing of place field widths in fear discrimination. Our head-fixed paradigm enables the recording of dendrites and the study of plasticity mechanisms (Sheffield et al., 2017; Sheffield & Dombeck, 2015), thus facilitating these investigations.

We recently used Paradigm 3 to understand the role of the thalamic nucleus reuniens-hippocampal pathway (Ratigan et al., 2023). Using a two-photon microscope, we identified neural activity in this pathway by imaging axons from the nucleus reuniens that project to the CA1 with single-axon resolution. We found this pathway necessary for suppressing fear memory retrieval, context discrimination, and fear extinction. The robust freezing response in Paradigm 3 in the fear-conditioned VR, compared to a neutral novel VR, allowed us to align our findings with the existing literature on the role of the nucleus reuniens-hippocampal circuit in fear memory in freely moving animals (Ramanathan & Maren, 2019; Ramanathan et al., 2018; Totty et al., 2023; Troyner et al., 2018; Vetere et al., 2017; Xu & Südhof, 2013). Furthermore, the subcellular resolution achieved through two-photon calcium imaging using the head-fixed preparation allowed us to longitudinally observe and understand the inputs that axons from the reuniens send to the hippocampus during fear memory formation and retrieval, a level of detail that is difficult to achieve with current techniques in freely moving preparations.

In another unpublished study, we used Paradigm 2 and identified co-active neurons between CA1 and CA3 subregions of the hippocampus during fear memory recall. The difference between the first and subsequent laps has helped us understand the progression of neural activity in the hippocampus as mice go from retrieving the fear memory to extinction within one session. Moreover, the variability in response across mice has allowed us to regress the neural activity to the behavior, identifying neural differences between those mice that accurately recall the fear memory and those that do not. These studies showcase the broad applicability of our different CFC paradigms.

In the laboratory, CFC provokes freezing behavior in freely moving rodents, as demonstrated by several studies that span decades (Michael S. Fanselow, 1980, 1990; Fendt & Fanselow, 1999; Maren, 2001; Maren et al., 2013; Phillips & LeDoux, 1992; Tovote et al., 2015). In our study, we found that a mild electric shock applied to the tail could also trigger freezing behavior as the conditioned response in head-fixed mice, making our paradigm more comparable to the freely moving paradigms than previously attempted methods using air-puff stimulation (Lovett-Barron et al., 2014; Rajasethupathy et al., 2015). There are numerous opportunities for future work to systematically test how different parameters—such as shock amplitude, number and spacing of shocks, duration of fear conditioning, context pre-exposure time, the configuration of Control versus CFC VR environments, and the inclusion of additional sensory cues—may influence both behavioral outcomes during recall and extinction, as well as the underlying neural representations (Michael S. Fanselow, 1990; Huckleberry et al., 2016; Ji & Maren, 2007; Maren & Holmes, 2016; Poulos et al., 2016). Our current study provides only an initial exploration of this parameter space.

In this manuscript, we focused on freezing as the primary conditioned response to align with studies of freely moving animals. However, we also see other defensive responses unique to our treadmill-based task. as well as fear-related hesitation and backward movements during memory recall—defensive behaviors that have also been observed in freely moving animals (Chu et al., 2024; Furuyama et al., 2023; Totty et al., 2021; Trott et al., 2022). Our technique enables us to conduct a more detailed analysis of the animal’s behavior. The head-fixed setup allows us to easily measure physiological responses including pupil diameter, respiration, vocalization, heart rate, and body temperature. These parameters are known to change with fear conditioning (Dupin et al., 2020; Godsil et al., 2000; Korn et al., 2017; Shionoya et al., 2013; Stiedl & Spiess, 1997). Such physiological measurements can also help differentiate between the various conditioned responses we observe. There are also opportunities to use machine-learning-empowered algorithms to parse out new behavior markers for contextual fear memory, such as gestures or facial expressions (Datta et al., 2019; Mathis & Mathis, 2020; Syeda et al., 2024; Wiltschko et al., 2020). Thus, we can understand the animal’s behavior in a well-controlled environment, which when combined with neural activity recording, can lead to new insights into the neural circuits mediating different defensive responses and emotional states.

Significant insights into memory mechanisms through the CFC task have also come from recent discoveries of memory engrams in the brain (Josselyn et al., 2015; Josselyn & Tonegawa, 2020; Liu et al., 2012; Tonegawa et al., 2018). Engram cells, identified in various brain regions with the help of genetic tagging and optogenetics, play a crucial role in encoding, consolidating, and retrieving specific memories. Typically, these cells are identified based on the expression of the immediate early gene Fos during fear conditioning. Their artificial activation with optogenetics can trigger memory recall, evidenced by increased freezing behavior. Understanding which neurons form these engram cells, how Fos expression is driven during memory formation, how memory information is stored in these cells, how they evolve with learning, and how their activation results in memory recall is an active area of research (Frankland et al., 2024; Mocle et al., 2024; Monasterio et al., 2024; Pettit et al., 2022; Suthard et al., 2024; Uytiepo et al., 2024). Incorporating our paradigm with simultaneous measurements of cellular and subcellular activity can help answer some of these questions. For instance, a recent study (Monasterio et al., 2024) tagged engram cells during fear conditioning while animals were freely moving and then imaged the spontaneous dynamics of these cells using two-photon imaging before and after fear conditioning. The study revealed that cells forming engrams were inherently more active. A limitation of this study was the separate contexts for imaging versus tagging, which can be addressed with our paradigm. Our CFC paradigm allows for direct neural recording of these cells before, during, and after tagging in the same context to examine how activity during memory formation relates to engram cell emergence, population dynamics, and memory recall. Furthermore, holographic stimulation techniques can reactivate specific cell ensembles of interest to help understand the causal relationships between engram cells and memory. Moreover, the hippocampus contains both place cells and engram cells, both of which appear to be involved in contextual memory. However, their relationship is not yet fully understood. Our CFC paradigm offers an opportunity to study both place cells, as we have shown here, and engram cells simultaneously. This can be achieved by incorporating techniques that allow for in vivo labeling of Fos (Pettit et al., 2022). Thus, our head-fixed version of the CFC paradigm opens new avenues for research into the neural underpinnings of fear learning and memory.

Another advantage of our method is that it allows for studying changes in neuronal dynamics in response to an aversive stimulus, such as a shock. Our fields of view during the shock and subsequent flight-like response periods remained stable, and we could correct image movements during these periods with Suite2p (Pachitariu et al., 2017). This is a significant improvement over one-photon and electrophysiological techniques, where abrupt movements such as those induced by an electric shock can change fields of view and reduce the likelihood of recording from the same groups of cells, thereby hindering longitudinal assessments of neural dynamics. In Paradigm 1, where we imaged hippocampal cells during shock, we did not find cells responding to the shock itself, as others have reported (Barth et al., 2023; Mocle et al., 2024; Suthard et al., 2024). We suspect this could be due to the number or duration of shocks administered and our shocks not being associated with a specific cue or location. In future studies, we plan to optimize these parameters to understand better how neural dynamics change in response to an aversive cue.

Finally, we recognize that our head-fixed CFC paradigm has some limitations. Compared to the freely moving CFC, our setup is less naturalistic. We defined different contexts as two visually distinct VR environments while odor, sound, and tactile cues remained constant. This could have made it harder for mice to differentiate between the two contexts, particularly in Paradigm 2. Unlike freely moving animals in CFC studies, our mice were water-restricted throughout the experimental process. Because the animals need to get used to wearing a tailcoat and running in the VRs without a reward, we tend to ’over-train’ the mice. This criterion excludes mice that are less adept at running, limiting the number that go through the process and requiring larger cohorts of mice. This also presents scalability challenges compared to freely moving paradigms, as successful implementation of the head-fixed CFC approach requires weeks of reward-based training, larger sample sizes, and a technically demanding VR setup that integrates multiple hardware and software components to precisely monitor behavior. Additionally, while rewards were removed during experimental sessions, water rewards were still used to motivate running during training. This reward-associated training could subtly influence later behavior. For instance, the increased freezing in the Familiar VR on later recall days and remapping of place cells in that context could reflect the lingering effects of reward-associated training. While Paradigms 2 and 3 address many of these concerns, our approach still differs from freely moving paradigms due to this initial reward-based training. The impact of using rewards during training on recall behavior needs further investigation. Furthermore, despite our best efforts to include equal numbers of male and female mice, only some female mice reached the experimental stage. This could be due to the tail-coat weight and the reduced weight of females after water restriction compared to males. Female mice of the same age (10-12 weeks) weighed significantly less than males (20-23 g vs 25-30 g) and lost further weight following water restriction. Although we observed no behavioral differences between the four tested females and the male mice, we cannot guarantee similar results in other female animals without further testing. Studies indicate that different sexes may display varying conditioned responses, as seen in rats (Gruene et al., 2015; S. Maren et al., 1994). In the future, we aim to address these limitations.

In conclusion, we have developed a head-fixed version of the classic CFC paradigm. Our findings demonstrate that mice exhibit freezing behavior when fear-conditioned in a VR environment. Moreover, we observed that hippocampal place cells remap and narrow their fields in response to fear conditioning, mirroring observations in freely moving animals. This innovative paradigm bridges the behavioral task gap for techniques requiring head-fixation, paving the way for new investigations into neural circuits and deepening our understanding of the neural foundations of emotional memory.

Methods

Subjects

All experimental and surgical procedures adhered to the University of Chicago Animal Care and Use Committee guidelines. This study utilized 10-12 week-old C57BL/6J, Thy1GCaMP6s and Tg(Grik4-cre)G32-4Stl/J mice. Following the start of water restriction, the mice were individually housed in a reverse 12-hour light/dark cycle, and behavioral experiments were conducted during the animals’ dark cycle.

Head-fixation surgery

Mice were anesthetized with 1-2% isoflurane and given an intraperitoneal injection of 0.5 mL of saline and a subcutaneous injection of 0.45 mL of Meloxicam (1-2 mg/kg). A small incision was made on the skull to attach a metal head-plate (9.1 mm x 31.7 mm, Atlas Tool and Die Works) to head-restrain the mice. Some mice also had an imaging window implanted above the hippocampus, as previously described (Dong et al., 2021; Krishnan et al., 2022; Sheffield & Dombeck, 2015). Post-surgery mice were housed individually. After a recovery day, they were water-restricted to 0.8-1.0 ml/day.

Behavior training

Head-restrained mice were trained to run on a cylindrical treadmill to navigate virtual reality (VR) environments similar to the setups previously described (Dong et al., 2021; Krishnan et al., 2022; Sheffield & Dombeck, 2015). Behavioral training began seven days following the water restriction. Briefly, VR environments were created using the Virtual Reality MATLAB Engine, or VIRMEN (Aronov & Tank, 2014), and projected to five screens covering the mouse’s field of view. Mice were head restrained with their limbs resting on a freely rotating Styrofoam wheel, which acted as a treadmill. VR environments were 2 m linear tracks. Mice received a water reward (4 μl) through a waterspout upon completing each traversal of the linear track, considered a lap. This reward was followed by a 1.5-second pause in the VR to allow the mouse to consume the water. The mice were then virtually teleported to the start of the track to begin a new lap. Training sessions lasted 30 minutes per day. Mice were considered well-trained when they could run at >3 laps per minute. We found that this level of training was necessary to ensure that mice would continue to run even without rewards (Krishnan et al., 2022; Krishnan & Sheffield, 2023). Generally, about 60% of mice met our criteria, which typically took around 10-14 days to achieve. Behavior data was collected using a PicoScope Oscilloscope (PICO4824, Pico Technology, v6.13.2).

Tail-coat

A custom-designed “tail-coat” delivered mild electric shocks to a head-fixed mouse’s tail during virtual reality (VR) navigation. A pictorial representation of the tail-coat can be found in Figure 1D. The tail-coat, handmade from wearable conductive fabrics, featured a lightweight design to ensure the mouse could continue running during baseline VR exploration (Table 7). First, the mouse’s tail was measured while anesthetized for head-plate surgeries. A small denim patch cloth (Michael’s) was cut into a rectangle to fit snugly on the mouse’s tail. Marks were made where snap buttons (Michael’s) would be sewn, ensuring they clipped just above the tail. Given that the mouse’s tail narrows away from the body, the buttons closer to the body were placed slightly further apart than those closer to the tip. Typically, the buttons were set 2 cm apart on the narrow end and 2.5 cm apart on the broader end. Care was taken to adjust the placement of the buttons so they weren’t too tight or loose for the mouse. We found that these measurements were generally similar across sexes and for the age group that we tested. After measurements were taken and the cloth cut, small strips of 8mm wide conductive fabric tape (Adafruit) were attached to either side of the cloth, with some tape hanging off the edge. A metal strip (McMaster Carr) of the same width was affixed to this portion of the tape to provide a secure grip for alligator clips used in fear conditioning. Finally, snap buttons were sewn with conductive thread at the marked spots. Clear nail polish was applied to secure the threads in place. The tail-coat creation process should take less than half an hour. Dimensions of the cloth for a typical male mouse, 12 weeks of age and weighing ∼30 g, are shown in Figure 1D. The denim cloth was 4 cm in width and 3.2 cm in length; the conductive tape was 1.2 cm in width and 5.2 cm in length, with 1.5 cm hanging out at each side.

Contextual fear conditioning protocol in head-fixed mice

Once the mice were well-trained and met our criteria of running >3 laps per minute, they were ready for the next stage. The following day, a tail-coat was attached to the mouse’s tail. Mice that maintained consistent running in the VR, with minimal pauses, while wearing the tail-coat advanced to the next stage. For mice startled by the tail-coat, we typically attempted another day with the tail-coat, rewarding them and observing if their running behavior improved to running at least three laps per minute. If not, these mice were not advanced to the next stage. In mice that do advance, on the subsequent day, the tail-coat was kept on, and the water reward was removed. All else remained the same, including the 1.5 s pause at the end, to allow animals to distinguish the start and end of a lap. If the mice continued to run as before, they moved on to the experiment stage. From that day forward, the mice no longer received water reinforcement when running in the VR environments. Typically, about 40% of mice made it to the experiment stage. Tail-coats were only added during VR exploration and removed before returning the animals to their home cage. Once the mice advanced to the experiment stage, they were taken through the entire process and not excluded from analysis, even if they showed signs of inconsistent running behavior, which happened with some mice (for instance, see Supplementary Figure 3B-C).

We implemented three distinct paradigms to assess contextual fear discrimination. In the first paradigm, we conducted experiments over two days (Paradigm 1). On Day 0, mice engaged with the same VR in which they were trained, now referred to as the Familiar VR. After 10 minutes, we transitioned them to a new VR for another 10 minutes; this new VR would later be associated with shocks (CFC VR). During this phase, the mice explored both VRs without any rewards while wearing the tail-coat. After the initial exploration in the CFC VR, they received six mild electric shocks, each lasting 1 second with 1-minute intervals in between. The shocks were administered using a shock generator (Colbourn Precision animal shocker, Harvard Bioscience). A TTL pulse was generated using custom-written codes in MATLAB to turn on the shock generator via an Arduino. The shock times were recorded by the PicoScope Oscilloscope (PICO4824, Pico Technology, v6.13.2). We tested different shock intensities ranging from 0.5 to 1.2 mA. After the last shock, mice were removed from the VR and returned to their home cage. The following day (Day 1), we placed the mice in the Familiar VR or CFC VR for ten minutes to test their memory recall. No rewards were delivered during this phase; however, the tail-coat was not added during memory recall. Some mice underwent additional recall tests over subsequent days to assess the time required for fear memory extinction. As a control, some mice underwent the same paradigm but without any shocks.

In the second paradigm (Paradigm 2), the experiment spanned over three days. On Day - 1, mice explored two novel VRs for ten minutes each, establishing a baseline for context exploration. One of these VRs served as a control, while the other, the CFC VR, was where the mice would later receive electric shocks. On Day 0, the mice again spent ten minutes each in the Control and CFC VRs. After this, they received six mild electric shocks, each lasting one second, with a strength of 1 mA and an inter-trial interval of one minute. The mice wore the tail-coat on both Days -1 and 0. On Day 1, the mice were placed in either the Control VR or CFC VR for ten minutes each, in a counterbalanced manner, to test for memory recall. No rewards were given on all the experimental days; however, the tail-coat was not added to the tails on the recall day (Day 1). The third paradigm (Paradigm 3) was similar to Paradigm 2 and used two novel VRs, except that a tail-coat was also added to the mice’s tails during recall tests.

Additionally, a subset of mice in Paradigms 1 and 2 and all mice in Paradigm 3 went through multiple days of recall tests. The subset of mice was randomly chosen, not based on their recall behavior from the previous day. Mice were presented with either the CFC VR or the Control VR in a counterbalanced manner.

All VRs were randomly assigned as either Control or CFC VR to prevent biasing any inherent preferences in mice. However, the Familiar VR was always kept the same. We also used both sexes in our experiments. However, we found female mice less receptive to the “tail-coat” than their male counterparts. We believe this is due to the low average weight of female mice (∼16 g) following water restriction. Female mice of the same age (10-12 weeks) weighed significantly less than males (20-23 g vs 25-30 g) and lost further weight following water restriction. As a result, only four female mice could proceed through and complete all the behavioral tests. Our ultimate goal is to compare behavior and neural activity across sexes, considering sex as a biological variable. To achieve this, we plan to adjust the age and use female mice of appropriate weight to ensure they can run on the treadmill with the tailcoat in the future.

Behavior measurement

Freezing periods were periods during which the animal’s instantaneous velocity was 0 cm/s. Freezing epochs were defined as continuous, uninterrupted freezing periods longer than 1 s. The time taken to complete a lap was calculated as the total time (in seconds) the animal took to run from 0 to 200 cm.

Surgery and two-photon imaging

Stereotaxic surgeries and cannula window implantation above the CA1 subregion of the hippocampus were performed in the same way as previously reported (Krishnan et al., 2022). Briefly, a genetically-encoded calcium indicator, AAV1-CamKII-GCaMP6f (pENN.AAV.CamKII.GCaMP6f.WPRE.SV40 was a gift from James M. Wilson – Addgene viral prep #100834-AAV1; https://www.addgene.org/100834/; RRID: Addgene_100834) was injected (∼50 nL at a depth of 1.25 mm below the surface of the dura) using a beveled glass micropipette leading to GCaMP6f expression in a large population of CA1 pyramidal cells. Mice underwent water restriction (1.0 ml/day) following viral injection and, seven days later, were implanted with a hippocampal window and head plate (Dombeck et al., 2010). Mice were allowed to rest from the surgeries for 3-4 days, after which behavioral training began. Imaging was done using a laser scanning two-photon microscope (Neurolabware). The setup is the same as the one previously described (Krishnan et al., 2022). Time-series images were collected through Scanbox (v4.1, Neurolabware), and the PicoScope Oscilloscope (PICO4824, Pico Technology, v6.13.2) was used to synchronize frame acquisition timing with behavior. At the end of the imaging session on Day -1 or Day 0, a 1-minute time-series movie was collected at a higher magnification and then averaged to aid as a reference frame in finding the same imaging plane on subsequent days. Across-day images collected were motion-corrected using Suite2p (Pachitariu et al., 2017) and aligned using ImageJ (v1.53, NIH). Region of interests were extracted using Suite2p.

Place cell extraction and place field parameters

Place fields were identified as previously described (Dong et al., 2021; Krishnan et al., 2022). Briefly, the 2 m track was divided into 40 position bins (each 5 cm wide). The running behavior of the animal was filtered to exclude periods where the animal was immobile (speed <1 cm/s). Filtering ensured that place cells were defined only during active exploration and not during freezing bouts. We have previously found that filtering versus not filtering yields similar results (Krishnan et al., 2022). Extracted place fields satisfied the following criteria, which were used for all conditions and all mice: 1. The average ΔF/F was greater than 10% above the baseline. 2. The cell displayed significant calcium transients on >30% of laps in the field. 3. The rising phase of the mean transient was located on the track. 4. Their p-value from bootstrapping was <0.05. Multiple place fields within the same cell were treated independently. Parameters of place fields were calculated as described before (Krishnan et al., 2022); this includes the center of mass (COM), place field reliability, out/in-field firing ratio, and place field widths.

Statistics

For related samples, we performed a paired t-test. Multiple comparisons of related samples were corrected with Bonferroni post-hoc. To quantify freezing across days, we employed a linear mixed-effects model with days of recall as the fixed effect and behavior prior to fear conditioning on Day 0 as baseline. The model also included a random effect for mice to account for the nesting of observations within individual subjects. Separate models were used for each VR. Boxplots are plotted to display the entire distribution of the data. The box in the boxplot ranges from the first quartile (25th percentile) to the third quartile (75th percentile), and the box shows the interquartile range (IQR). The line across the box represents the median (50th percentile). The whiskers extend to 1.5*IQR on either side of the box, and anything above this range is defined as an outlier. A two-tailed Kolmogorov-Smirnov (KS) Test was used to compare distributions. When not using boxplots, mean and confidence intervals are displayed.